---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# isw

`isw` prepares stream networks and models stream depletion and aquifer
water-level change caused by groundwater pumping. Its main workflow separates
physical stream segments from pump-specific apportionment and supports both
analytical depletion-function (ADF) and constant-head stream injection.

## Installation

Install the development version from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("gopalpenny/isw")
```

## Example

This example uses the spatial data included with the package.

```{r example, message=FALSE}
library(isw)
library(units)

stream_segments <- prep_stream_segments(
  example_stream_reaches,
  reach_spacing = set_units(200, "m")
)

stream_apportionment <- prep_adf_stream_apportionment(
  example_pumping_wells,
  stream_segments,
  sample_spacing = set_units(50, "m")
)

pumping_schedules <- tibble::tibble(
  t = set_units(c(0, 30, 60), "days"),
  pump_1 = set_units(c(400, 400, 0), "m^3/day"),
  pump_2 = set_units(c(0, 250, 0), "m^3/day")
)

evaluation_times <- set_units(c(0, 30, 60, 90), "days")

stream_depletion <- model_adf_stream_depletion(
  example_pumping_wells,
  pumping_schedules,
  stream_apportionment,
  evaluation_times
)

water_level_change <- model_aquifer_water_level_change(
  example_pumping_wells,
  pumping_schedules,
  example_observation_wells,
  stream_segments,
  evaluation_times,
  stream_apportionment = stream_apportionment
)
```

The map shows the total modeled stream-depletion rate at day 60. Pumping wells
are red triangles and observation wells are black points.

```{r example-figure, message=FALSE, fig.width=7, fig.height=5}
depletion_map <- stream_depletion |>
  dplyr::filter(evaluation_time == set_units(60, "days")) |>
  dplyr::group_by(reach_id, reach_segment_id) |>
  dplyr::summarize(
    depletion_m3_day = as.numeric(set_units(
      sum(stream_depletion_rate),
      "m^3/day"
    )),
    .groups = "drop"
  )

plot_segments <- dplyr::left_join(
  stream_segments,
  depletion_map,
  by = c("reach_id", "reach_segment_id")
)

ggplot2::ggplot() +
  ggplot2::geom_sf(
    data = plot_segments,
    ggplot2::aes(color = depletion_m3_day),
    linewidth = 2
  ) +
  ggplot2::geom_sf(
    data = example_pumping_wells,
    color = "firebrick",
    shape = 17,
    size = 3
  ) +
  ggplot2::geom_sf(
    data = example_observation_wells,
    color = "black",
    size = 2.5
  ) +
  ggplot2::labs(color = "Stream depletion\n(m3/day)") +
  ggplot2::theme_minimal()
```

See the [package workflow](https://gopalpenny.github.io/isw/articles/stream-depletion-and-drawdown.html)
for ADF and constant-head examples, and the
[function reference](https://gopalpenny.github.io/isw/reference/) for complete
input and output descriptions.
