---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# isw

<!-- badges: start -->
<!-- badges: end -->

The goal of isw is to enable modeling of stream depletion and aquifer drawdown.

## Installation

You can install the development version of isw from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("gopalpenny/isw")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(isw)
```

Consider the following configuration of stream, pumping well, and observation well.

```{r fig.width = 3, fig.height = 2, echo = FALSE}
library(png)
well_config_img <- readPNG("fig/pumping_observation_wells.png")
# grid::grid.raster(well_config_img)

p_well_config <- ggplot2::ggplot() + ggplot2::annotation_custom(grid::rasterGrob(well_config_img,
                                                                width=ggplot2::unit(1,"npc"),
                                                                height=ggplot2::unit(1,"npc")),
                                               -Inf, Inf, -Inf, Inf) + ggplot2::coord_equal()
p_well_config
```


Stream depletion and aquifer drawdown in this scenario can be modeled using the function
`get_depletion_from_pumping` in this package. 

```{r message = FALSE}
library(units)
x1 <- set_units(c(1, 5, 10) * 1e3, "ft")
x2 <- set_units(1e3, "ft")
y <- set_units(1e3, "ft")
D <- set_units(100, "ft")
K <- set_units(0.001, "ft/sec")
t <- set_units(5, "year")
V <- 0.2 # unitless
get_depletion_from_pumping(x1, x2, y, K, D, V, t)
```

For details of this function, check `?get_depletion_from_pumping`.

