---
title: "Getting started with ISW package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ISW package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette illustrates key elements of the `isw` package, used to generate
simple models of interconnected surface water. The package includes functions to
estimate stream depletion and aquifer drawdown.

The idealized model in `isw` assumes the wells are screened over the saturated thickness of the aquifer, the stream fully penetrates the aquifer, all flow is horizontal, and the transmissivity is uniform and insensitive to changes in water level.

The package can be installed using `remotes::install_github("gopalpenny/isw")`, and loaded
as:

```{r setup, message=FALSE}
library(isw)
```

## Main package functions

The `isw` package has three primary functions to understand interconnected surface waters:

* `stream_depletion_fraction`: Estimate stream depletion due to pumping well(s)
* `aquifer_drawdown_ratio`: Estimate aquifer drawdown due to pumping well(s) (no stream)
* `get_depletion_from_pumping`: Estimate stream depletion and aquifer drawdown

We'll see how each of these works, below. Note that within these models, representation
of the stream uses the [method of well images](https://gopalpenny.github.io/anem/articles/anem-vignette.html),
which basically says that streams with constant head can be modeled by mirroring
wells across the stream and switching the sign of pumping. In other words, a pumping well and
and injection well on opposite sides of the stream will ensure constant hydraulic head
at the midpoint between the two wells, assuming the rates of pumping and injection
are equal in magnitude. 

Note that the `aquifer_drawdown_ratio` does not assume a stream, and therefore
works only for a single well. Both `stream_depletion_fraction` and `get_depletion_from_pumping`
account for an image well on the other side of the stream.

## Aquifer parameters 

Aquifers in `isw` are represented with just a few variables. Let's consider an 
aquifer with the following characteristics:

- Saturated thickness (`D`): 100 ft
- Permeability (`K`): 0.001 ft/sec
- Drainable porosity of the aquifer (`V`): 0.2

Note that `isw` uses the `units` package to keep track of dimensionality, 
therefore we can begin by loading this package. The `set_units` function can be 
used to specify units objects, and therefore aquifer parameters.

```{r message = FALSE}
library(units)
D <- set_units(100, "ft")
K <- set_units(0.001, "ft/sec")
t <- set_units(5, "year")
V <- 0.2 # unitless
```

## Some basic calculations

### Estimating stream depletion with `stream_depletion_fraction()`

The `stream_depletion_fraction()` function estimates stream depletion at time `t` as a
fraction of pumping from an individual pumping well, following Glover and Balmer (1954). The
`x1` parameter represents distance from the stream. We can call `stream_depletion_fraction`
by specifying `x1`, the aquifer parameters, and the elapsed time `t` since pumping initiated.

```{r}
x1 <- set_units(0.5, "mi")
stream_depletion_fraction <- get_stream_depletion_fraction(x1 = x1, K = K, D = D, V = V, t = t)
stream_depletion_fraction
```

We can get total stream depletion by multiplying this fraction by a pumping rate,
for instance 1 cfs:

```{r}
pumping_rate <- set_units(1, "ft^3/s")
stream_depletion <- stream_depletion_fraction * pumping_rate
stream_depletion
```

### Estimating aquifer drawdown with `get_aquifer_drawdown_ratio()` 

The `get_aquifer_drawdown_ratio()` function gives the change in water level per unit pumping
at an observation well (without a stream). The parameter `r` is the distance between 
the pumping well and observation well.

```{r}
r <- set_units(1, "mi")
aquifer_drawdown_ratio <- get_aquifer_drawdown_ratio(r = r, K = K, D = D, V = V, t = t)
aquifer_drawdown_ratio
```

Note that `get_aquifer_drawdown_ratio` has an optional parameter, `well_diam` that defaults to a value of 0. When specified, drawdown does not increase inside the well. In other words, when `r<well_diam/2`, drawdown is calculated as if `r=well_diam/2`.

### Depletion and drawdown with `get_depletion_from_pumping()`

The `get_depletion_from_pumping()` function estimates stream depletion fraction (using
`get_stream_depletion_fraction`) and changes in water level at an
observation well (`get_aquifer_drawdown_ratio`) due to abstraction from a
pumping well at time `t` after pumping initiates. Like
`get_stream_depletion_fraction`, and unlike `get_aquifer_drawdown_ratio`,
The function accounts for a the effect of the stream as a constant head
boundary using, as noted above, the method of images.

This function uses a couple new parameters. The `x2` parameter ($x_2$) is the distance 
between the observation well and the stream. If `r` is the distance between
the pumping and observation wells, `y` is the component of this distance that is
parallel to the stream. In other words, $r^2 = (x_2-x_1)^2 + y^2$. Keeping the remaining
parameters the same as before, we can call this function as:

```{r}
x2 <- set_units(1e3, "ft")
y <- set_units(1e3, "ft")
depletion_from_pumping <- get_depletion_from_pumping(x1 = x1,
                                                     x2 = x2,
                                                     y = y,
                                                     K = K,
                                                     D = D,
                                                     V = V,
                                                     t = t)
depletion_from_pumping
```

## Parameter combinations with `vector` and `data.frame` objects

We can also estimate drawdown at multiple distances from a pumping well. We use the 
function `get_aquifer_drawdown_ratio`, noting that both `stream_depletion_fraction`
and `get_depletion_from_pumping` also allow multiple parameter combinations specified 
as `vector` or `data.frame` objects. Let's get this ratio a distances ranging from 
1 ft to 10 miles, after 2 years of pumping.

```{r}
r <- set_units(seq(1/5280, 10, length.out = 200), "mi")
t <- set_units(2, 'year')
aquifer_drawdown_ratio <- get_aquifer_drawdown_ratio(r = r, K = K, D = D, V = V, t = t)
head(aquifer_drawdown_ratio)
```

Equivalently, we can specify parameter combinations in a `data.frame`. The `tidyr`
function `crossing()` can be very helpful to get multiple parameter combinations.

```{r}
library(tidyr)

# create tibble with all parameter combinations
df <- tibble(K = K, D = D, V = V, t = t) %>%
  crossing(r = set_units(seq(1/5280, 10, length.out = 200), "mi"))

# call function with tibble to specify parameters
aquifer_drawdown_ratio <- get_aquifer_drawdown_ratio(df)
head(aquifer_drawdown_ratio)
```


For simplicity, let's assume the well was pumping 1 cusec over those two years. The change in water level is calculated as `aquifer_drawdown_ratio` multiplied by the pumping rate.

```{r}
pumping_rate <- set_units(1, "ft^3/sec")
aquifer_drawdown <- aquifer_drawdown_ratio * pumping_rate
```

We can then plot the drawdown as a function of pumping using `ggplot.` The `bind_cols`
function from `dplyr` allows us to join the drawdown results with the original `data.frame`.

```{r message =FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
drawdown_df <- df %>% 
  bind_cols(drawdown = aquifer_drawdown)

ggplot(drawdown_df, aes(r, drawdown)) +
  geom_line()
```

## Checking the drawdown calculations

As a check, to give confidence in this calculation, we can calculate the volume of water drained from the aquifer as drawdown times the drainable porosity for circular bands (with differential radius $dr$) around the well. For simplicity, we use the `dplyr::mutate` function to calculate this value.

```{r message = FALSE}
dr <- r[2] - r[1]
drawdown_df <- drawdown_df %>% 
  mutate(d_area = dr * 2 * pi * r,
         volumetric_drawdown = drawdown * d_area * V)
```

In an aquifer with no recharge, the total volumetric drawdown should be equal to the pumping rate over the time period, 1 cusec over five years or `r round(pumping_rate * t/1e6, 0)` million ft^3^. The water drained from the aquifer can be calculated using `sum(volumetric_drawdown)`, which gives a result of `r round(sum(set_units(drawdown_df$volumetric_drawdown,"ft^3"))/1e6,0)` million ft^3^. 

The total radial flow towards the well can be calculated as $q = 2 \pi r D K \frac{\partial h}{\partial r}$ (the typical negative sign in Darcy's law is removed because we are calculating flow in the direction of negative r). We can run this calculation and plot flow towards the aquifer.

```{r}
drawdown_df <- drawdown_df %>% 
  mutate(flow_towards_well = 2 * pi * r * D * K * (drawdown - lag(drawdown)) / dr,
         flow_towards_well = set_units(flow_towards_well, "ft^3/s"))

ggplot(drawdown_df, aes(r, flow_towards_well)) + geom_line() +
  geom_abline(slope = 0, intercept = 1, color = "red", linetype = "dashed")
```

Note that closer to the pumping well this calculation suffers from coarse discretization of the domain causing the flowrate to diverge from 1 ft^3^/s. You can verify this yourself by, for instance, changing the radius to ` r <- set_units(seq(1, 5, length.out = 200), "ft")` and re-running the calculations.

## References

Glover, Robert E., and Glenn G. Balmer. “River Depletion Resulting from Pumping a Well near a River.” Eos, Transactions American Geophysical Union 35, no. 3 (1954): 468–70. https://doi.org/10.1029/TR035i003p00468.

