---
title: "Identify-blind-spots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Identify-blind-spots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(isw)
```



```{r}
library(patchwork)
library(units)
library(png)
library(dplyr)
library(ggplot2)
library(tidyr)
```

The package handles the configuration between river,
pumping well, and observation well in a particular way, as shown in the figure below. 
We need to keep track of the following variables:

- `x1`: Distance from the pumping well to the river
- `x2`: Distance from the observation well to the river
- `y`: Component of the distance between the two wells that is *parallel* to the river


```{r echo = FALSE}
library(png)
well_config_img <- readPNG("../inst/figs/pumping_observation_wells.png")

p_well_config <- ggplot() + annotation_custom(grid::rasterGrob(well_config_img,
                                                                width=unit(1,"npc"),
                                                                height=unit(1,"npc")),
                                               -Inf, Inf, -Inf, Inf) +
  coord_equal()
p_well_config
```

We also need to consider aquifer parameters. We parameterize an aquifer with the 
following characteristics, based on the Glover paper:

- Saturated thickness (`D`): 100 ft
- Permeability (`K`): 0.001 ft/sec
- Drainable porosity of the aquifer (`V`): 0.2

```{r}
# example from Glover
# x1 <- set_units(c(1, 5, 10, 25, 50) * 1e3, 'ft')
D <- set_units(100, 'ft')
K <- set_units(0.001, 'ft/sec')
# t_sec <- set_units(1, "year") # sec in a year
V <- 0.2
```

## Prepare themes for plotting

To make the plots look nice, we can create the following themes.

```{r}
txt_size <- 16
theme_basic <- theme(
  line = element_line(size = 1, color = "black"), 
  panel.background = element_rect(fill = "white"), 
  legend.text = element_text(size = txt_size*0.9, color = "black"), 
  axis.title.x = element_text(size = txt_size, colour = "black"), 
  axis.title.y = element_text(size = txt_size, angle = 90, colour = "black"), 
  legend.title = element_text(size = txt_size, color = "black"))


theme_std <- theme_basic %+replace% theme(
  panel.border = element_rect(fill = NA, colour = "black", size = 1), 
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), axis.line = element_blank(), 
  axis.ticks.x = element_line(colour = "black", size = 0.5), 
  axis.ticks.y = element_line(colour = "black", size = 0.5), 
  axis.ticks.length = unit(1, "mm"), 
  axis.text.x = element_text(size = txt_size*0.9, colour = "black"), 
  axis.text.y = element_text(size = txt_size*0.9,  colour = "black", hjust = 1))

theme_diagram <- theme_basic %+replace% 
  theme(
  panel.border = element_rect(fill = NA, colour = NA, size = 1), 
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  axis.line = element_blank(), 
  axis.ticks.x = element_blank() , 
  axis.ticks.y = element_blank(), 
  axis.ticks.length = unit(1, "mm"), 
  axis.text.x = element_blank(), 
  axis.text.y = element_blank(), 
  axis.title.x = element_blank(), 
  axis.title.y = element_blank()) %+replace%
  theme(legend.position = c(0.8, 0.75), legend.box = 'vertical', legend.title = element_blank())
```



## Set up grid of locations

We first generate a `tibble` of different permutations of distances among the 
river, pumping well, and observation well.

```{r}
# Pumping well locations
x1_coords <- seq(-1, 6, by = 0.05)
y1_coords <- seq(-1, 11, by = 0.2)
# x1_coords <- seq(0, 5, by = 0.05)
# y1_coords <- seq(0, 10, by = 0.2)
pump_wells <- crossing(x1 = x1_coords, y1 = y1_coords) %>%
  mutate(keep = x1 > 0 & x1 <= 5 & y1 >= 0 & y1 <=10,
         across(all_of(c("x1","y1")), function(x) set_units(x,"mi")))

# Observation well locations
x2_coords <- c(0.8, 1.6, 1, 3, 4)
y2_coords <- c(2, 4, 8.5, 1, 6)
well_diam <- set_units(50, "ft")

# Times
t_years <- 5 #c(0.05, seq(0.1,5, by = 0.1))
  
obs_wells <- tibble(x2 = x2_coords, y2 = y2_coords, well_diam) %>%
  mutate(`Obs. well` = as.factor(row_number()),
         across(all_of(c("x2","y2")), function(x) set_units(x,"mi")))

wells <- crossing(pump_wells,
                  obs_wells,
                  t = set_units(t_years,'year')) %>%
  mutate(K = K, D = D, V = V,
         y = abs(y2 - y1),
         id = row_number())



# locations <- tribble(~x1, ~y, ~`Obs. well`, ~`Pumping well`,
#                      0.2, 1.5, "a", "c",
#                      0.8, 1.5, "a", "d",
#                      0.2, 0, "b", "c",
#                      0.8, 0, "b", "d") %>%
#   mutate(x2 = 0.5,
#          across(all_of(c("x1","x2","y")), function(x) set_units(x,"mi")), id = row_number())
```

To visualize these permutations we can plot them.

```{r}
p_grid <- ggplot() + 
  geom_point(data = pump_wells, aes(x1, y1), size = 0.5) + 
  geom_point(data = pump_wells %>% filter(!keep), aes(x1, y1), size = 0.5, color = "red") + 
  geom_point(data = obs_wells, aes(x2, y2, shape = `Obs. well`, color = `Obs. well`), size = 4) + 
  # scale_shape_manual(values = c(0, 5)) +
  xlab("Distance from river (x)") + 
  ylab("Distance along river (y)") +
  # scale_color_discrete("Well type") + 
  theme_std +
  coord_equal()
p_grid
```

Now we can get aquifer depletion with the `get_depletion_from_pumping` function. As seen 
in the output, the function returns a `data.frame` with two columns:

- `stream_depletion_fraction`: The fraction of pumping at that timestep that is
associated with stream depletion (opposed to drawing down the aquifer)
- `aquifer_drawdown_ratio`: The amount of aquifer drawdown (i.e., change in groundwater level) 
per unit pumping. 

```{r}
depletion <- get_depletion_from_pumping(wells %>% mutate(x1=abs(x1)))
head(depletion)
```

We can now add the depletion to the tibble to the original data frame for plotting and analysis.

```{r}
df_abline <- crossing(slope = c(1,2,4), intercept = 0)
wells_calcs <- wells %>%
  bind_cols(as_tibble(depletion)) %>%
  mutate(ratio = -stream_depletion_fraction / aquifer_drawdown_ratio,
         ratio_cut = cut(ratio, c(0, df_abline$slope, Inf)))
head(wells_calcs)
```


```{r}
p_ratio <- ggplot(wells_calcs %>% group_by(x1, y1) %>% filter(ratio == min(ratio), keep)) + 
  geom_point(aes(as.numeric(abs(aquifer_drawdown_ratio)), as.numeric(stream_depletion_fraction), alpha = as.numeric(x1), color = ratio_cut)) +
  geom_abline(data = df_abline, aes(slope = slope, intercept = intercept)) +
  geom_abline(data = df_abline[df_abline$slope==1,], aes(slope = slope, intercept = intercept),color="red",size=1) +
  scale_color_viridis_d("Stream depletion (cfs)\nper aquifer drawdown (ft)")+
  scale_alpha_continuous("Distance\nfrom river",range = c(1, 0.2)) +
  guides(color = "none") +
  theme_std +
  scale_x_continuous("Aquifer drawdown per cfs pumping (ft)", limits=c(0, 3)) + 
  scale_y_continuous("Stream depletion per cfs pumping (cfs)", limits = c(0,1), breaks = seq(0,1,by = 0.2))
p_ratio
```

Now let's plot the same ratio as a map

```{r}
wells_calcs_ratio <- wells_calcs %>% 
  group_by(x1, y1) %>% filter(ratio == min(ratio)) %>%
  mutate(ratio = as.numeric(ratio),
         blind_spot = as.numeric(ratio) > 2)
```


```{r}
p_blind_spot<- ggplot(wells_calcs_ratio %>% group_by(x1, y1) %>% filter(ratio == min(ratio))) + 
  # geom_contour_filled(aes(x1, y1, z = as.numeric(ratio)), breaks = c(0,1,2,4, Inf)) +
  geom_contour_filled(aes(x1, y1, z = as.numeric(ratio)), breaks = c(0,1,2,4, Inf), color = "black") +
  geom_contour(aes(x1, y1, z = as.numeric(ratio)), breaks = c(1,2), color = "red") +
  geom_contour(aes(x1, y1, z = as.numeric(ratio)), breaks = c(0,2,4, Inf), color = "black") +
  geom_point(data = obs_wells, aes(x2, y2, shape = `Obs. well`, color = `Obs. well`), size = 4) + 
  scale_fill_viridis_d("Stream depletion (cfs)\nper aquifer drawdown (ft)") +
  scale_x_units("Distance from river") +
  ylab("Distance along river") +
  theme_std +#%+replace% 
  # theme(
  #   panel.margin = unit(c(0, 0, 0, 0), "null"),
  #   plot.margin = unit(c(0, 0, 0, 0), "null")) +
  coord_equal(xlim = c(0,5),ylim = c(0,10),expand=FALSE)
p_blind_spot
```

```{r fig.width=14,fig.height=6}
p_ratio + p_blind_spot
```

