---
title: "Stream Depletion Apportionment: Detailed Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stream Depletion Apportionment: Detailed Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages}
library(isw)
library(units)
library(tibble)
library(dplyr)
library(ggplot2)
library(sf)
```

## Purpose and scope

This vignette follows the internal steps used to apportion stream depletion
and calculate aquifer drawdown. It complements the public workflow in
`vignette("stream-depletion-apportionment")`.

Functions called with `isw:::` are implementation details. They are shown to
make the calculation transparent, but they may change without the compatibility
guarantees associated with exported functions. Application code should
normally call the exported functions demonstrated in the main vignette.

## Inputs

The walkthrough uses the package's synthetic spatial inputs. There are two
pumping wells, two observation wells, and three reaches that meet at a
confluence.

```{r spatial-inputs}
pumping_wells <- example_pumping_wells
stream_reaches <- example_stream_reaches
observation_wells <- example_observation_wells

ggplot() +
  geom_sf(data = stream_reaches, color = "steelblue", linewidth = 1) +
  geom_sf(data = pumping_wells, color = "firebrick", size = 3) +
  geom_sf(data = observation_wells, color = "black", size = 3) +
  theme_minimal()
```

The pumping schedule contains a constant interval, a rate reduction, and a
shutdown for `pump_1`. `pump_2` starts later and then shuts down. Evaluation
times are deliberately independent of these schedule boundaries.

```{r temporal-inputs}
pumping_schedules <- tibble(
  t = as.Date(c("2025-01-01", "2025-02-01", "2025-03-01", "2025-04-01")),
  pump_1 = set_units(c(500, 500, 300, 0), "m^3/day"),
  pump_2 = set_units(c(0, 250, 250, 0), "m^3/day")
)

evaluation_times <- seq.Date(
  from = as.Date("2025-01-16"),
  by = "15 days",
  length.out = 8
)
```

## Validate and prepare spatial inputs

The public modeling functions first validate identifiers, geometry types,
aquifer properties, physical units, and coordinate reference systems. The
analysis requires projected coordinates because its distances must have
linear units.

`.select_analysis_crs()` finds a local UTM CRS from the combined spatial
extent unless the user supplies another projected CRS. `.prepare_spatial_inputs()`
then transforms copies of every object to that CRS, drops vertical coordinate
dimensions, and supplies a zero-meter well diameter when none is given.

```{r prepare-spatial-inputs}
analysis_crs <- isw:::.select_analysis_crs(
  pumping_wells,
  stream_reaches,
  observation_wells
)

spatial_inputs <- isw:::.prepare_spatial_inputs(
  pumping_wells,
  stream_reaches,
  observation_wells,
  analysis_crs = analysis_crs
)

spatial_inputs$analysis_crs$Name
spatial_inputs$pumping_wells$well_diam
```

## Discretize the stream network

`.discretize_stream_reaches()` divides each input reach into approximately
equal segments whose lengths do not exceed `reach_spacing`. The active
geometry remains the segment line. A second geometry column, `model_point`,
stores the along-line midpoint later used as the location of the equivalent
stream-injection well.

```{r discretize-stream-reaches}
reach_segments <- isw:::.discretize_stream_reaches(
  spatial_inputs$stream_reaches,
  reach_spacing = set_units(100, "m")
)

reach_segments[c(
  "reach_id",
  "reach_segment_id",
  "represented_length"
)]
```

```{r plot-reach-segments}
ggplot() +
  geom_sf(data = spatial_inputs$stream_reaches, color = "grey75", linewidth = 3) +
  geom_sf(data = reach_segments, aes(color = reach_segment_id), linewidth = 1.5) +
  geom_sf(
    data = reach_segments,
    aes(geometry = model_point),
    shape = 21,
    fill = "white",
    color = "black",
    size = 2.5
  ) +
  geom_sf(data = spatial_inputs$pumping_wells, color = "firebrick", size = 3) +
  labs(color = "Reach segment") +
  theme_minimal()
```

## Sample each reach segment

`sample_reach_segments()` represents the shape of each segment with points at
the centers of equal-length intervals. `sampled_length` records how much
stream length each point represents. This finer sampling geometry is used only
to calculate spatial weights; it is distinct from the single `model_point`
used in the drawdown model.

```{r sample-reach-segments}
sample_points <- sample_reach_segments(
  reach_segments,
  sample_spacing = set_units(25, "m")
)

sample_points[c(
  "reach_id",
  "reach_segment_id",
  "sample_point_id",
  "sampled_length"
)]
```

```{r plot-sample-points}
ggplot() +
  geom_sf(data = reach_segments, color = "grey65", linewidth = 1.5) +
  geom_sf(data = sample_points, color = "royalblue", size = 1.5) +
  geom_sf(
    data = reach_segments,
    aes(geometry = model_point),
    shape = 21,
    fill = "white",
    color = "black",
    size = 3
  ) +
  geom_sf(data = spatial_inputs$pumping_wells, color = "firebrick", size = 3) +
  theme_minimal()
```

## Calculate static reach apportionment

For each pump and sample point, web-squared apportionment is proportional to
the point's represented length divided by squared pump-to-point distance.
Point weights are normalized within each pump and then summed by reach
segment. Zero-distance points receive special handling so division by zero
cannot occur.

`get_stream_reach_apportionment()` performs the preparation, discretization,
sampling, weighting, and aggregation steps together. It also calculates the
shortest distance from each pump to each segment line. That segment distance,
rather than the sample-point distance, is supplied to the analytical response
kernel.

```{r stream-apportionment}
stream_apportionment <- get_stream_reach_apportionment(
  pumping_wells,
  stream_reaches,
  reach_spacing = set_units(100, "m"),
  sample_spacing = set_units(25, "m"),
  method = "web_squared"
)

stream_apportionment |>
  st_drop_geometry() |>
  group_by(pump_id) |>
  summarize(
    number_of_segments = n(),
    total_fraction = sum(apportionment_fraction),
    .groups = "drop"
  )
```

## Convert schedules to pumping events

Analytical response functions are linear, so intermittent pumping is handled
by superposition of changes in pumping rate. `.get_pumping_rate_changes()`
compares each scheduled rate with the preceding rate, assuming zero pumping
before the first row. Exact zero changes are omitted.

```{r pumping-rate-changes}
pumping_rate_changes <- isw:::.get_pumping_rate_changes(
  pumping_schedules,
  pumping_wells
)

pumping_rate_changes
```

`.get_pumping_response_times()` pairs every nonzero change with evaluation
times strictly later than that event. It also calculates the positive elapsed
time used by the analytical kernel. An event beginning exactly at an
evaluation boundary has not yet produced a response at that boundary.

```{r pumping-response-times}
pumping_response_times <- isw:::.get_pumping_response_times(
  pumping_schedules,
  pumping_wells,
  evaluation_times
)

head(pumping_response_times)
```

## Evaluate analytical depletion fractions

The Glover response depends on pump-to-stream distance, hydraulic diffusivity,
and elapsed time. `.get_stream_depletion_fraction_lookup()` evaluates each
unique pump, segment, and elapsed-time combination once. Repeated event rows
can then reuse the lookup rather than recalculating the same analytical value.

```{r depletion-fraction-lookup}
depletion_fraction_lookup <- isw:::.get_stream_depletion_fraction_lookup(
  pumping_wells,
  pumping_response_times,
  stream_apportionment
)

head(depletion_fraction_lookup)
```

For each event and segment, the package multiplies the signed pumping-rate
change by the analytical depletion fraction and the static apportionment
fraction. Event contributions are then summed by pump, segment, and evaluation
time. The exported function performs these joins and restores explicit zeroes
for combinations without a contributing event.

```{r apportioned-stream-depletion}
stream_depletion <- get_apportioned_stream_depletion(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  evaluation_times
)

head(stream_depletion)
```

## Construct the injection time grid

Stream recovery is approximated with piecewise-constant injection at each
segment model point. `.get_stream_injection_times()` combines pumping
boundaries, evaluation times, and any optional refinement times. Pumping
boundaries are retained even when output is requested less frequently.

```{r injection-time-grid}
injection_times <- seq.Date(
  from = pumping_schedules$t[1],
  to = max(evaluation_times),
  by = "10 days"
)

injection_grid <- isw:::.get_stream_injection_times(
  pumping_schedules,
  evaluation_times,
  injection_times
)

injection_grid$injection_times
```

Depletion is evaluated at all injection boundaries. The negative arithmetic
mean of adjacent endpoint rates becomes the constant injection rate for that
interval. `.get_injection_rate_changes()` then converts consecutive interval
rates to change events so the same superposition machinery can be used for
injection and pumping.

```{r internal-injection-schedule}
depletion_at_injection_times <- get_apportioned_stream_depletion(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  injection_grid$injection_times
)

internal_injection_schedule <-
  isw:::.get_interval_average_injection_schedule(
    depletion_at_injection_times,
    pumping_schedules
  )

injection_rate_changes <- isw:::.get_injection_rate_changes(
  internal_injection_schedule
)

head(internal_injection_schedule)
head(injection_rate_changes)
```

The public function performs these steps and converts the internal elapsed-day
boundaries back to the user's original time representation.

```{r public-injection-schedule}
stream_injection_schedule <- get_stream_injection_schedule(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  evaluation_times,
  injection_times = injection_times
)

head(stream_injection_schedule)
```

## Superimpose pumping drawdown and stream recovery

`get_apportioned_aquifer_drawdown()` calculates the direct Theis response from
each physical pumping well to each observation well. It separately calculates
the response from every segment model point using the injection-rate changes.
Both calculations use the aquifer properties of the originating pump.

Pumping drawdown and stream recovery are reported as positive magnitudes. Net
aquifer drawdown is their difference. Results remain separated by pump and
observation well so users can inspect contributions before aggregating them.

```{r apportioned-aquifer-drawdown}
aquifer_drawdown <- get_apportioned_aquifer_drawdown(
  pumping_wells,
  pumping_schedules,
  observation_wells,
  stream_apportionment,
  evaluation_times,
  stream_injection_schedule = stream_injection_schedule
)

head(aquifer_drawdown)
```

```{r plot-aquifer-drawdown}
aquifer_drawdown |>
  select(
    pump_id,
    observation_id,
    evaluation_time,
    pumping_drawdown,
    stream_recovery,
    aquifer_drawdown
  ) |>
  tidyr::pivot_longer(
    cols = c(pumping_drawdown, stream_recovery, aquifer_drawdown),
    names_to = "component",
    values_to = "water_level"
  ) |>
  ggplot(aes(evaluation_time, water_level, color = component)) +
  geom_line() +
  facet_grid(observation_id ~ pump_id) +
  labs(
    x = "Evaluation time",
    y = "Water-level response",
    color = "Component"
  ) +
  theme_minimal()
```
