---
title: "Stream Depletion and Drawdown in the ISW package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stream Depletion and Drawdown in the ISW package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  out.width = "100%",
  fig.align = "center"
)
```

```{r packages, message = FALSE}
library(isw)
library(units)
library(tibble)
library(dplyr)
library(ggplot2)
library(sf)
```

## Overview

This vignette presents the main `isw` workflow for estimating stream
depletion and the resulting water-level changes in an aquifer. It compares
the package's two methods for estimating depletion and stream-injection 
using spatial inputs included with the package:

* `method = "adf"`, the default, calculates analytical depletion and
  apportions it among stream segments; and
* `method = "constant_head"` solves the injection rates needed to keep
  water-level change equal to zero at the stream-segment model points.

The focus here is how to prepare the inputs and connect the main functions.
For more detail, see the website articles on the
[comparison of the two methods](https://gopalpenny.github.io/isw/articles/stream-injection-methods.html)
and the
[internals of ADF apportionment](https://gopalpenny.github.io/isw/articles/stream-depletion-apportionment-internals.html).

## Load the spatial inputs

The package includes a synthetic example domain with two pumping wells, three
connected stream reaches, and two observation wells, all as `sf` objects.

```{r spatial-inputs}
pumping_wells <- example_pumping_wells
stream_reaches <- example_stream_reaches
observation_wells <- example_observation_wells

ggplot() +
  geom_sf(data = stream_reaches, color = "grey65", linewidth = 1.2) +
  geom_sf(data = pumping_wells, color = "firebrick", shape = 8, size = 3) +
  geom_sf(data = observation_wells, color = "#0072B2", shape = 17, size = 3) +
  geom_sf_text(data = pumping_wells, aes(label = pump_id), nudge_y = 100) +
  geom_sf_text(
    data = observation_wells,
    aes(label = observation_id),
    nudge_y = -100
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal()
```

See the
[`get_stream_segments()` reference](https://gopalpenny.github.io/isw/reference/get_stream_segments.html)
for CRS requirements, coordinate transformations, and automatic selection of
a projected analysis CRS.

Required identifiers for spatial inputs are `reach_id`, `pump_id`, and `observation_id`, respectively.
Pumping wells must also provide aquifer properties as `units` vectors including: 
hydraulic conductivity `K`, aquifer thickness `D`, and
drainable porosity `V`. The optional `well_diam` prevents a point-well
singularity when a response is evaluated inside the well radius.

The individual objects and their columns can be inspected directly.

```{r inspect-spatial-inputs}
pumping_wells
stream_reaches
observation_wells
```

## Divide the stream into model segments

Both the `adf` and `constant-head` methods can use the same stream segments object. 
`reach_spacing` is the
maximum modeled segment length; smaller values represent the stream boundary
at greater spatial resolution but increase the number of injection rates that
must be calculated.

Each returned segment contains a `model_point`. This is the discrete injection
location and, for constant head, the location where the zero-change condition
is imposed. The effective injection-well diameter is based on the segment's
`represented_length`.

```{r stream-segments}
stream_segments <- get_stream_segments(
  stream_reaches,
  reach_spacing = set_units(400, "m")
)

stream_segments[c(
  "reach_id",
  "reach_segment_id",
  "represented_length",
  "well_diam"
)]
```

The same reference describes the returned columns in more detail.

## Define pumping and model times

`pumping_schedules` is a wide tibble, with pumping provided in units. 
Its first column, `t`, gives the start of
each pumping period; every other column is named for a `pump_id` and contains
that well's rate. Rates continue until the next schedule row, so a shutdown
must be represented by a row that sets the rate to zero.

```{r pumping-schedule}
pumping_schedules <- tibble(
  t = as.Date(c("2025-01-01", "2025-02-01", "2025-03-01", "2025-04-01")),
  pump_1 = set_units(c(500, 500, 300, 0), "m^3/day"),
  pump_2 = set_units(c(0, 250, 250, 0), "m^3/day")
)

pumping_schedules
```

`evaluation_times` specifies when stream depletion and water-level results are
returned. Pumping-schedule times and evaluation times are always included as
boundaries of the piecewise-constant stream-injection schedule.

The optional `injection_times` argument can add schedule boundaries without
requesting additional water-level output. This can improve the interval-average
approximation to continuously changing ADF depletion, but it increases
computation for both methods. Constant-head rates are solved to satisfy the
boundary at each interval end. Very short intervals can therefore require large
corrective rates, especially when a regular refinement grid is combined with
nearby evaluation or pumping times. This example leaves `injection_times` as
`NULL` and uses monthly evaluation boundaries for a stable, readily interpreted
comparison. Applications that add refinement times should inspect sensitivity
to the resulting interval lengths.

```{r model-times}
evaluation_times <- seq.Date(
  from = as.Date("2025-01-01"),
  to = as.Date("2026-01-01"),
  by = "month"
)

injection_times <- NULL
```

See the
[`get_stream_injection_schedule()` reference](https://gopalpenny.github.io/isw/reference/get_stream_injection_schedule.html)
for the complete input requirements and interval construction rules.

## Construct both stream-injection schedules

ADF requires one additional step because its total analytical depletion must
be apportioned spatially. `get_adf_stream_apportionment()` calculates a set of
segment weights independently for each pumping well. `sample_spacing`
controls how finely curved stream geometry is sampled when those weights are
calculated; it does not change the number of modeled stream segments.

```{r adf-apportionment}
stream_apportionment <- get_adf_stream_apportionment(
  pumping_wells,
  stream_segments,
  sample_spacing = set_units(100, "m"),
  method = "web_squared"
)

stream_apportionment |>
  st_drop_geometry() |>
  group_by(pump_id) |>
  summarize(
    total_apportionment_fraction = sum(apportionment_fraction),
    .groups = "drop"
  )
```

We can first calculate ADF stream depletion directly at the requested times.
This is the simplest route when only ADF depletion, rather than its effect on
aquifer water levels, is needed.

```{r direct-adf-depletion}
adf_depletion <- get_adf_stream_depletion(
  pumping_wells = pumping_wells,
  pumping_schedules = pumping_schedules,
  stream_apportionment = stream_apportionment,
  evaluation_times = c(evaluation_times, injection_times)
)

adf_depletion_totals <- adf_depletion |>
  st_drop_geometry() |>
  group_by(pump_id, evaluation_time) |>
  summarize(
    adf_depletion_at_time = sum(stream_depletion_rate),
    .groups = "drop"
  )

adf_depletion_totals
```

The two interval schedules can now be constructed with the same function and
time grid. `"adf"` is the default method, but it is written explicitly here to
make the comparison clear.

With `method = "adf"`, `get_stream_injection_schedule()` calls
`get_adf_stream_depletion()` at every schedule boundary and averages depletion
at adjacent boundaries. It reports the resulting aquifer injection as a signed
`injection_rate`, which is negative when water enters the aquifer. Its magnitude
is the average stream-depletion rate applied over the interval. By contrast,
`get_adf_stream_depletion()` returns positive depletion at each requested time.
If ADF depletion has already been calculated at every schedule boundary, it can
be supplied through `adf_stream_depletion` to avoid recalculation. The supplied
result must include all pumping, evaluation, and optional refinement boundaries
used by the schedule.

With `method = "constant_head"`, each returned interval rate is the rate
solved to satisfy the constant-head boundary condition at the interval end.

```{r injection-schedules}
adf_schedule <- get_stream_injection_schedule(
  pumping_wells = pumping_wells,
  pumping_schedules = pumping_schedules,
  stream_segments = stream_segments,
  evaluation_times = evaluation_times,
  injection_times = injection_times,
  method = "adf",
  stream_apportionment = stream_apportionment,
  adf_stream_depletion = adf_depletion
)

constant_head_schedule <- get_stream_injection_schedule(
  pumping_wells = pumping_wells,
  pumping_schedules = pumping_schedules,
  stream_segments = stream_segments,
  evaluation_times = evaluation_times,
  injection_times = injection_times,
  method = "constant_head"
)
```

The constant-head result includes two diagnostics. `boundary_residual` is the
signed water-level change remaining at a segment model point after the solve.
`matrix_condition_number` indicates the numerical sensitivity of the system;
`isw` warns when it exceeds \(10^6\).

An aquifer factorization is the decomposition of the stream-response matrix
used to solve for the constant-head depletion rates. Wells with the same
aquifer properties share this matrix, so `isw` factorizes it once for that
aquifer-property combination and reuses the result for each associated pumping
well. This reduces computation while preserving the depletion response
attributed to each pump.

```{r constant-head-diagnostics}
constant_head_schedule |>
  st_drop_geometry() |>
  summarize(
    maximum_absolute_boundary_residual_m = max(abs(as.numeric(
      set_units(boundary_residual, "m")
    ))),
    maximum_matrix_condition_number = max(matrix_condition_number),
    number_of_aquifer_factorizations = n_distinct(aquifer_id)
  )
```

## Compare stream depletion

`injection_rate` is negative when water leaves the stream and enters the
aquifer. For each pumping well, total stream depletion is therefore the
negative of the injection rate summed across all stream segments.

For ADF, the schedule contains interval-average analytical depletion. For
constant head, it contains the depletion required to impose the stream
boundary condition at the interval end.

```{r summarize-stream-depletion}
stream_depletion <- bind_rows(
  mutate(adf_schedule, method = "ADF"),
  mutate(constant_head_schedule, method = "Constant head")
) |>
  st_drop_geometry() |>
  group_by(method, pump_id, interval_start, interval_end) |>
  summarize(
    stream_depletion_rate = -sum(injection_rate),
    .groups = "drop"
  )

stream_depletion |>
  filter(interval_end %in% evaluation_times)
```

The direct ADF result at an interval end generally differs from the average
rate applied during that interval. The following comparison makes that
distinction explicit.

```{r compare-adf-endpoint-and-average}
adf_interval_comparison <- adf_schedule |>
  st_drop_geometry() |>
  group_by(pump_id, interval_start, interval_end) |>
  summarize(
    adf_interval_average = -sum(injection_rate),
    .groups = "drop"
  ) |>
  left_join(
    rename(
      adf_depletion_totals,
      adf_depletion_at_interval_end = adf_depletion_at_time
    ),
    by = c(
      "pump_id",
      "interval_end" = "evaluation_time"
    )
  )

adf_interval_comparison
```

```{r plot-stream-depletion}
adf_depletion_plot <- adf_depletion_totals |>
  transmute(
    pump_id,
    interval_end = evaluation_time,
    stream_depletion_rate = adf_depletion_at_time,
    method = "ADF at requested time"
  )

ggplot() +
  geom_step(
    data = stream_depletion,
    aes(interval_start, stream_depletion_rate, color = method),
    linewidth = 1
  ) +
  geom_line(
    data = adf_depletion_plot,
    aes(interval_end, stream_depletion_rate, color = method),
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_point(
    data = adf_depletion_plot,
    aes(interval_end, stream_depletion_rate, color = method),
    size = 2
  ) +
  facet_wrap(~pump_id, nrow = 1) +
  labs(
    x = "Interval end",
    y = "Total stream-depletion rate",
    color = "Method"
  ) +
  theme_minimal()
```

The following map compares the spatial distribution of the two injection
schedules during May 2025. Because package injection rates are negative when
water enters the aquifer, the mapped values are their positive magnitudes.
Each panel shows the rates attributed to one pumping well and one method.

```{r map-may-injection-rates, fig.height = 9}
may_date <- as.Date("2025-05-15")

may_injection_rates <- bind_rows(
  mutate(adf_schedule, method = "ADF"),
  mutate(constant_head_schedule, method = "Constant head")
) |>
  filter(interval_start <= may_date, interval_end > may_date) |>
  group_by(method, pump_id, reach_segment_id) |>
  summarize(
    injection_rate = sum(injection_rate),
    .groups = "drop"
  ) |>
  mutate(
    injection_rate_m3_day = -as.numeric(set_units(
      injection_rate,
      "m^3/day"
    ))
  )

injection_points <- st_sf(
  reach_segment_id = stream_segments$reach_segment_id,
  geometry = stream_segments$model_point
) |>
  left_join(may_injection_rates, by = "reach_segment_id")

ggplot() +
  geom_sf(data = stream_segments, color = "grey75", linewidth = 1) +
  geom_sf(
    data = injection_points,
    aes(
      size = injection_rate_m3_day,
      color = injection_rate_m3_day
    ),
    alpha = 0.85
  ) +
  facet_grid(method ~ pump_id) +
  scale_size_continuous(range = c(2, 10)) +
  scale_color_viridis_c(option = "C") +
  labs(
    x = NULL,
    y = NULL,
    size = "Injection rate\n(m³/day)",
    color = "Injection rate\n(m³/day)"
  ) +
  theme_minimal()
```

## Calculate aquifer drawdown and stream recovery

`get_aquifer_water_level_change()` combines the effects of the physical
pumping wells and stream injection at each observation well and evaluation
time. Supplying a previously calculated schedule ensures that the water-level
calculation uses exactly the injection rates inspected above.

```{r calculate-water-levels}
adf_water_levels <- get_aquifer_water_level_change(
  pumping_wells = pumping_wells,
  pumping_schedules = pumping_schedules,
  observation_wells = observation_wells,
  stream_segments = stream_segments,
  evaluation_times = evaluation_times,
  stream_injection_schedule = adf_schedule
) |>
  mutate(method = "ADF")

constant_head_water_levels <- get_aquifer_water_level_change(
  pumping_wells = pumping_wells,
  pumping_schedules = pumping_schedules,
  observation_wells = observation_wells,
  stream_segments = stream_segments,
  evaluation_times = evaluation_times,
  stream_injection_schedule = constant_head_schedule
) |>
  mutate(method = "Constant head")

water_levels <- bind_rows(adf_water_levels, constant_head_water_levels)

water_levels |>
  st_drop_geometry() |>
  select(
    method,
    pump_id,
    observation_id,
    evaluation_time,
    pumping_drawdown,
    stream_recovery,
    water_level_change
  )
```

`pumping_drawdown` and `stream_recovery` are reported as positive magnitudes.
`water_level_change` is their signed net effect: negative values indicate a
decline and positive values indicate a rise.

```{r plot-water-levels, fig.height = 8}
water_level_totals <- water_levels |>
  st_drop_geometry() |>
  group_by(method, pump_id, observation_id, evaluation_time) |>
  summarize(
    water_level_change_m = as.numeric(set_units(
      sum(water_level_change),
      "m"
    )),
    .groups = "drop"
  )

ggplot(
  water_level_totals,
  aes(evaluation_time, water_level_change_m, color = method)
) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_line(linewidth = 1, alpha = 0.8) +
  facet_grid(observation_id ~ pump_id, scales = "fixed") +
  labs(
    x = "Evaluation time",
    y = "Net water-level change (m)",
    color = "Method"
  ) +
  theme_minimal()
```

The
[`get_aquifer_water_level_change()` reference](https://gopalpenny.github.io/isw/reference/get_aquifer_water_level_change.html)
documents the returned pumping, recovery, and net water-level components. The
[method-comparison article](https://gopalpenny.github.io/isw/articles/stream-injection-methods.html)
extends this workflow with observations along the stream and on transects,
and compares the discrete constant-head result with the analytical image-well
solution.

## Map water-level change along the stream

The stream boundary can also be evaluated directly by treating points along
the stream as observation wells. Here, the segment `model_point` locations are
combined with more closely spaced points generated along each segment. The
model points are the locations where the constant-head condition is imposed, and
the additional sample points show how water levels behave between them.

```{r stream-observation-points}
stream_sample_points <- generate_segment_sample_points(
  stream_segments,
  sample_spacing = set_units(100, "m")
) |>
  transmute(
    observation_id = paste0("sample_", sample_point_id),
    point_type = "Sample point"
  )

stream_model_points <- st_sf(
  observation_id = paste0(
    "model_",
    stream_segments$reach_segment_id
  ),
  point_type = "Segment model point",
  geometry = stream_segments$model_point
)

stream_observation_wells <- bind_rows(
  stream_sample_points,
  stream_model_points
)
```

In the figure below, negative values indicate a decline from
the initial condition. The following calculation sums the two pump-specific
responses at each observation point on May 1, 2025.

```{r calculate-stream-water-levels}
stream_map_time <- as.Date("2025-05-01")

adf_stream_water_levels <- get_aquifer_water_level_change(
  pumping_wells = pumping_wells,
  pumping_schedules = pumping_schedules,
  observation_wells = stream_observation_wells,
  stream_segments = stream_segments,
  evaluation_times = stream_map_time,
  stream_injection_schedule = adf_schedule
) |>
  mutate(method = "ADF")

constant_head_stream_water_levels <- get_aquifer_water_level_change(
  pumping_wells = pumping_wells,
  pumping_schedules = pumping_schedules,
  observation_wells = stream_observation_wells,
  stream_segments = stream_segments,
  evaluation_times = stream_map_time,
  stream_injection_schedule = constant_head_schedule
) |>
  mutate(method = "Constant head")

stream_water_level_map <- bind_rows(
  adf_stream_water_levels,
  constant_head_stream_water_levels
) |>
  st_drop_geometry() |>
  group_by(method, observation_id, evaluation_time) |>
  summarize(
    water_level_change_m = as.numeric(set_units(
      sum(water_level_change),
      "m"
    )),
    .groups = "drop"
  ) |>
  left_join(
    stream_observation_wells,
    by = "observation_id"
  ) |>
  st_as_sf()
```

```{r map-stream-water-levels, fig.height = 10}
water_level_limit <- max(abs(
  stream_water_level_map$water_level_change_m
))

ggplot() +
  geom_sf(data = stream_segments, color = "grey75", linewidth = 1) +
  geom_sf(
    data = stream_water_level_map,
    aes(
      fill = water_level_change_m,
      shape = point_type
    ),
    color = "black",
    size = 4.5,
    stroke = 0.35
  ) +
  facet_wrap(~method, ncol = 1) +
  scale_fill_gradient2(
    low = "#B2182B",
    mid = "white",
    high = "#2166AC",
    midpoint = 0,
    limits = c(-water_level_limit, water_level_limit)
  ) +
  scale_shape_manual(values = c(
    "Sample point" = 21,
    "Segment model point" = 23
  )) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Net water-level\nchange (m)",
    shape = "Observation type"
  ) +
  theme_minimal()
```

Finally, we can summarize the distribution of hydraulic-head change by method
and observation type to confirm that the constant-head method is working at the segment model points.

```{r summarize-stream-water-levels}
stream_water_level_summary <- stream_water_level_map |>
  st_drop_geometry() |>
  group_by(method, point_type) |>
  summarize(
    mean_m = mean(water_level_change_m),
    standard_deviation_m = sd(water_level_change_m),
    minimum_m = min(water_level_change_m),
    maximum_m = max(water_level_change_m),
    .groups = "drop"
  )

stream_water_level_summary
```
