---
title: "Stream Depletion Apportionment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stream Depletion Apportionment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages}
library(isw)
library(units)
library(tibble)
library(dplyr)
library(ggplot2)
library(sf)
```

## Introduction

This vignette develops a stream-depletion apportionment example from the
package's spatial and pumping-schedule inputs. We begin with the inputs and
functions that are currently available. At this stage, depletion is calculated
using the distance to the closest stream reach. Later sections will distribute
that depletion among the discretized stream reaches.

## Define the pumping well

`pumping_wells` is an `sf` point object. Each pumping well has a unique
`pump_id` and the aquifer properties associated with that well. Aquifer
properties retain their physical units.

```{r pumping-wells}
pumping_wells <- st_as_sf(
  tibble(
    pump_id = "pump_1",
    x = 499800,
    y = 4980050,
    K = set_units(1e-5, "m/s"),
    D = set_units(50, "m"),
    V = 0.1
  ),
  coords = c("x", "y"),
  crs = 32615
)

pumping_wells
```

## Define the pumping schedule and evaluation times

`pumping_schedules` uses wide format. Its first column contains the schedule
times, and every remaining column is named using a `pump_id` from
`pumping_wells`. The pumping rate begins at its corresponding schedule time.

This first example uses one constant pumping event of 500 cubic meters per day.
We evaluate its response at the end of each of the following 100 days.

```{r pumping-schedule}
pumping_schedules <- tibble(
  t = as.Date("2025-01-01"),
  pump_1 = set_units(500, "m^3/day")
)

evaluation_times <- seq.Date(
  from = pumping_schedules$t + 1,
  by = "day",
  length.out = 100
)

pumping_schedules
head(evaluation_times)
```

## Define the stream network

`stream_reaches` is an `sf` line object with one row and one unique `reach_id`
for each input stream reach. Here, two upstream reaches meet at a confluence
and flow into a downstream reach.

```{r stream-reaches}
stream_reaches <- st_sf(
  reach_id = c("upstream_1", "upstream_2", "downstream"),
  geometry = st_sfc(
    st_linestring(
      matrix(c(500000, 4980200, 500150, 4980050), ncol = 2, byrow = TRUE)
    ),
    st_linestring(
      matrix(c(500000, 4979900, 500150, 4980050), ncol = 2, byrow = TRUE)
    ),
    st_linestring(
      matrix(c(500150, 4980050, 500350, 4979850), ncol = 2, byrow = TRUE)
    ),
    crs = 32615
  )
)

ggplot() +
  geom_sf(data = stream_reaches, linewidth = 1) +
  geom_sf(data = pumping_wells, color = "red", size = 3) +
  theme_minimal()
```

## Prepare the spatial inputs

The spatial inputs can be supplied in any defined coordinate reference system.
For distance-based calculations, they are transformed to one projected
analysis CRS. Because `well_diam` was not supplied, the prepared pumping-well
object assigns it a default value of zero meters.

The spatial-preparation function is currently internal. It will eventually be
called by the public stream-depletion interface.

```{r prepare-spatial-inputs}
spatial_inputs <- isw:::.prepare_spatial_inputs(
  pumping_wells,
  stream_reaches
)

spatial_inputs$analysis_crs$Name
spatial_inputs$pumping_wells$well_diam
```

## Discretize the stream network

Each input reach is divided into reach segments that do not exceed the requested
spacing. The active geometry remains a line segment, while `model_point`
contains the point at the center of each reach segment. `represented_length`
records the actual stream length represented by that point.

```{r discretize-stream-reaches}
reach_segments <- isw:::.discretize_stream_reaches(
  spatial_inputs$stream_reaches,
  reach_spacing = set_units(100, "m")
)

reach_segments[c(
  "reach_id",
  "reach_segment_id",
  "represented_length"
)]
```

```{r plot-discretized-stream-network}
ggplot() +
  geom_sf(
    data = spatial_inputs$stream_reaches,
    color = "grey75",
    linewidth = 3
  ) +
  geom_sf(
    data = reach_segments,
    aes(color = reach_segment_id),
    linewidth = 1.5
  ) +
  geom_sf(
    data = reach_segments,
    aes(geometry = model_point),
    color = "black",
    fill = "white",
    shape = 21,
    size = 2.5
  ) +
  geom_sf(
    data = spatial_inputs$pumping_wells,
    color = "red",
    size = 3
  ) +
  labs(color = "Reach segment") +
  theme_minimal()
```

## Prepare the pumping response times

The pumping schedule is converted into changes in pumping rate. Each nonzero
change is then paired with the evaluation times at which it can affect stream
depletion. `elapsed_time` is the time passed to the analytical response
function.

```{r prepare-pumping-response-times}
pumping_response_times <- isw:::.get_pumping_response_times(
  pumping_schedules,
  spatial_inputs$pumping_wells,
  evaluation_times
)

pumping_response_times
```

## Calculate depletion from the closest stream

Before adding apportionment, we can reproduce the conventional analytical
calculation by treating the closest reach as the single infinite stream. The
distance is calculated from the pumping well to the original stream geometry.

```{r closest-stream-distance}
pump_to_reach_distances <- st_distance(
  spatial_inputs$pumping_wells,
  spatial_inputs$stream_reaches
)

distance_to_stream <- min(pump_to_reach_distances)
distance_to_stream
```

The aquifer properties are joined to the pumping-event table using `pump_id`.
The descriptive `distance_to_stream` variable is passed to the existing Glover
kernel as its `x1` argument.

```{r calculate-stream-depletion}
df <- pumping_response_times |>
  left_join(
    st_drop_geometry(spatial_inputs$pumping_wells),
    by = "pump_id"
  ) |>
  mutate(
    distance_to_stream = distance_to_stream,
    stream_depletion_fraction = get_stream_depletion_fraction(
      x1 = distance_to_stream,
      K = K,
      D = D,
      V = V,
      t = elapsed_time
    ),
    stream_depletion_rate =
      stream_depletion_fraction * pumping_rate_change
  )

df |>
  select(
    pump_id,
    evaluation_time,
    elapsed_time,
    distance_to_stream,
    stream_depletion_fraction,
    stream_depletion_rate
  )
```

```{r plot-stream-depletion}
ggplot(df, aes(evaluation_time, stream_depletion_rate)) +
  geom_line() +
  labs(
    x = "Evaluation time",
    y = "Stream-depletion rate"
  ) +
  theme_minimal()
```

This calculation uses only the closest stream distance and does not distribute
depletion among the reach segments. We will retain it as a baseline for
comparison with the apportioned results.

## Sample points within each reach segment

Web apportionment represents the geometry of each reach segment with multiple
points. `sample_spacing` controls this numerical resolution independently of
the `reach_spacing` used to define the modeled reach segments.

```{r sample-reach-segments}
sample_points <- sample_reach_segments(
  reach_segments,
  sample_spacing = set_units(25, "m")
)

sample_points[c(
  "reach_id",
  "reach_segment_id",
  "sample_point_id",
  "sampled_length"
)]
```

```{r plot-reach-sample-points}
ggplot() +
  geom_sf(data = reach_segments, color = "grey65", linewidth = 1.5) +
  geom_sf(data = sample_points, color = "blue", size = 1.5) +
  geom_sf(
    data = reach_segments,
    aes(geometry = model_point),
    color = "black",
    fill = "white",
    shape = 21,
    size = 3
  ) +
  geom_sf(
    data = spatial_inputs$pumping_wells,
    color = "red",
    size = 3
  ) +
  theme_minimal()
```

The blue points are used to resolve geometry during apportionment. The larger
white points are the single model locations associated with each reach
segment.

## Calculate stream-depletion apportionment

The web-squared method weights each sample point using its represented stream
length and inverse squared distance from the pumping well. Point weights are
summed by `reach_segment_id` and normalized separately for every pump.

The same function also calculates `pump_to_reach_distance`, the shortest
distance from the pump to each reach-segment line. This segment-specific
distance is used by the analytical stream-depletion function.

```{r calculate-stream-apportionment}
stream_apportionment <- get_stream_depletion_apportionment(
  pumping_wells,
  stream_reaches,
  reach_spacing = set_units(100, "m"),
  sample_spacing = set_units(25, "m"),
  method = "web_squared"
)

stream_apportionment[c(
  "pump_id",
  "reach_id",
  "reach_segment_id",
  "pump_to_reach_distance",
  "apportionment_fraction"
)]
```

The apportionment fractions must sum to one within each pumping well:

```{r check-apportionment-balance}
stream_apportionment |>
  st_drop_geometry() |>
  group_by(pump_id) |>
  summarize(total_apportionment_fraction = sum(apportionment_fraction))
```

```{r plot-stream-apportionment}
ggplot(stream_apportionment) +
  geom_sf(aes(color = apportionment_fraction), linewidth = 2) +
  geom_sf(
    data = spatial_inputs$pumping_wells,
    color = "red",
    size = 3
  ) +
  labs(color = "Apportionment fraction") +
  theme_minimal()
```

## Apply the pumping schedule

We now apply the pumping schedule to every pump--reach-segment pair. The
analytical stream-depletion fraction is specific to `pump_id`,
`reach_segment_id`, and `elapsed_time`. Internally, the function calculates a
lookup table of unique elapsed-time fractions so repeated response periods do
not repeat the analytical calculation.

Each pumping-rate change is multiplied by its time-dependent analytical
fraction and the static apportionment fraction. Contributions from all earlier
pumping events are then summed using superposition.

```{r calculate-apportioned-stream-depletion}
stream_depletion <- get_apportioned_stream_depletion(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  evaluation_times
)

stream_depletion
```

The returned data remain separated by pumping well and reach segment. This
allows individual contributions to be inspected before combining them.

```{r plot-apportioned-stream-depletion}
ggplot(
  stream_depletion,
  aes(
    evaluation_time,
    stream_depletion_rate,
    color = reach_segment_id
  )
) +
  geom_line() +
  labs(
    x = "Evaluation time",
    y = "Stream-depletion rate",
    color = "Reach segment"
  ) +
  theme_minimal()
```

Total modeled depletion across the stream network can be calculated by
summing the reach-segment contributions:

```{r total-apportioned-stream-depletion}
total_stream_depletion <- stream_depletion |>
  group_by(evaluation_time) |>
  summarize(
    stream_depletion_rate = sum(stream_depletion_rate),
    .groups = "drop"
  )

total_stream_depletion
```

## Estimate aquifer drawdown and stream recovery

We now add an observation well. Aquifer drawdown is calculated using the
direct distance from the pumping well to the observation well, without an
image well. Each reach segment's `model_point` is then treated as an injection
well whose rate is derived from that segment's stream depletion.

```{r observation-wells}
observation_wells <- st_as_sf(
  tibble(
    observation_id = "obs_1",
    x = 499900,
    y = 4980100
  ),
  coords = c("x", "y"),
  crs = 32615
)

ggplot() +
  geom_sf(data = stream_reaches, color = "blue") +
  geom_sf(data = pumping_wells, color = "red", size = 3) +
  geom_sf(data = observation_wells, color = "black", size = 3) +
  theme_minimal()
```

The continuously changing stream-depletion rate is approximated over each
pumping-schedule interval using the mean of its values at the beginning and
end of the interval. The negative of that interval-average rate is applied as
injection at the segment model point. Evaluation times control when drawdown
is reported, not the internal injection timestep.

The injection schedule can be inspected before calculating drawdown.

```{r construct-stream-injection-schedule}
stream_injection_schedule <- get_stream_injection_schedule(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  evaluation_times
)

stream_injection_schedule
```

```{r calculate-apportioned-aquifer-drawdown}
aquifer_drawdown <- get_apportioned_aquifer_drawdown(
  pumping_wells,
  pumping_schedules,
  observation_wells,
  stream_apportionment,
  evaluation_times,
  stream_injection_schedule = stream_injection_schedule
)

aquifer_drawdown
```

The output separates the decline caused by the physical pumping well from the
recovery caused by stream injection. `aquifer_drawdown` is the pumping
drawdown minus stream recovery.

```{r plot-apportioned-aquifer-drawdown}
aquifer_drawdown |>
  select(
    evaluation_time,
    pumping_drawdown,
    stream_recovery,
    aquifer_drawdown
  ) |>
  tidyr::pivot_longer(
    cols = -evaluation_time,
    names_to = "component",
    values_to = "water_level"
  ) |>
  ggplot(aes(evaluation_time, water_level, color = component)) +
  geom_line() +
  labs(
    x = "Evaluation time",
    y = "Water-level response",
    color = "Component"
  ) +
  theme_minimal()
```
