---
title: "Stream Depletion Apportionment: Detailed Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stream Depletion Apportionment: Detailed Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  out.width = "100%",
  fig.align = "center"
)
```

```{r packages-isw}
library(isw)
```


```{r packages, message=FALSE}
library(units)
library(tibble)
library(dplyr)
library(ggplot2)
library(sf)
```

## Purpose and scope

This vignette repeats the workflow in
`vignette("stream-depletion-apportionment")` in the same order. After each
top-level ADF function call, it opens that function to show the intermediate
objects, calculations, and modeling choices involved.

Functions called with `isw:::` are implementation details. They are useful for
understanding and diagnosing a calculation, but they may change without the
compatibility guarantees associated with exported functions. Application code
should normally use the top-level functions shown at the start of each section.

This detailed walkthrough still names the combined compatibility functions
`get_stream_reach_apportionment()`, `get_apportioned_stream_depletion()`, and
`get_apportioned_aquifer_drawdown()` while explaining their internals. They are
planned for deprecation. New application code should use
`get_stream_segments()`, `get_adf_stream_apportionment()`,
`get_adf_stream_depletion()`, and `get_aquifer_water_level_change()`. See
`vignette("stream-injection-methods")` for the preferred shared workflow and
the constant-head alternative.

## Load the example spatial inputs

The walkthrough uses the same two pumping wells, three connected stream
reaches, and two observation wells as the main vignette.

```{r spatial-inputs}
pumping_wells <- example_pumping_wells
stream_reaches <- example_stream_reaches
observation_wells <- example_observation_wells

ggplot() +
  geom_sf(data = stream_reaches, color = "steelblue", linewidth = 1) +
  geom_sf(data = pumping_wells, color = "firebrick", size = 3) +
  geom_sf(data = observation_wells, color = "black", size = 3) +
  geom_sf_text(data = pumping_wells, aes(label = pump_id), nudge_y = 100) +
  geom_sf_text(
    data = observation_wells,
    aes(label = observation_id),
    nudge_y = -100
  ) +
  theme_minimal()
```

### What the spatial inputs control

`pumping_wells` supplies one point and one unique `pump_id` per physical well.
It also supplies the hydraulic conductivity `K`, aquifer thickness `D`,
drainable porosity `V`, and optional `well_diam` used by the analytical
responses. Here, `well_diam` is 0.3 m for both wells.

```{r inspect-pumping-wells}
pumping_wells
```

`stream_reaches` supplies the line geometry to which depletion will be
apportioned. The user supplies a unique `reach_id` for every input feature.
The package later creates `reach_segment_id` values when it discretizes those
features; users should not add that reserved column themselves.

```{r inspect-stream-reaches}
stream_reaches
```

`observation_wells` is needed only for aquifer drawdown. Each point has a
unique `observation_id`. All spatial inputs must have a defined CRS, but they
do not need to begin in the same CRS because the modeling functions transform
copies to a common projected analysis CRS.

## Define pumping schedules and evaluation times

The temporal inputs are identical to those in the main vignette. Schedule
times indicate changes in pumping rate; evaluation times indicate when output
is requested. Results are evaluated on the 1st and 15th of each month, plus
January 1, 2026 as the year-end boundary. This shows the post-shutdown response
across the full year while limiting vignette build time.

```{r pumping-schedule}
pumping_schedules <- tibble(
  t = as.Date(c("2025-01-01", "2025-02-01", "2025-03-01", "2025-04-01")),
  pump_1 = set_units(c(500, 500, 300, 0), "m^3/day"),
  pump_2 = set_units(c(0, 250, 250, 0), "m^3/day")
)

month_starts <- seq.Date(
  from = as.Date("2025-01-01"),
  to = as.Date("2025-12-01"),
  by = "month"
)
evaluation_times <- sort(c(
  month_starts,
  month_starts + 14,
  as.Date("2026-01-01")
))

pumping_schedules
```

### What happens to the temporal inputs

The schedule must be in wide format. Its first column is `t`, and every other
column name must exactly match one `pump_id`. Times must increase strictly,
and pumping rates must have flow units. A row's rates remain in effect until
the next row. The rates in the final row continue indefinitely; the package
does not assume that pumping stops after the final schedule time. To represent
a shutdown, users must include a final row that sets the applicable pumping
rates to zero, as this example does on 2025-04-01.

Internally, `Date` values and unit-bearing relative times are normalized to
elapsed days. The original representation is restored in the returned output.
Evaluation times are independent of the schedule and may be more or less
frequent. Passing `evaluation_times = NULL` to the modeling functions requests
output at `pumping_schedules$t`.

## Apportioning depletion among stream reaches

The first ADF modeling step assigns a static fraction of each pump's
eventual stream depletion to every modeled reach segment.

```{r calculate-stream-apportionment}
stream_apportionment <- get_stream_reach_apportionment(
  pumping_wells,
  stream_reaches,
  reach_spacing = set_units(400, "m"),
  sample_spacing = set_units(100, "m"),
  method = "web_squared"
)

stream_apportionment[c(
  "pump_id",
  "reach_id",
  "reach_segment_id",
  "pump_to_reach_distance",
  "apportionment_fraction"
)]
```

### What happens inside `get_stream_reach_apportionment()`

#### 1. Prepare a common analysis CRS

The function validates the wells and reaches, selects a projected analysis
CRS, transforms copies of the inputs, drops Z and M coordinates, and supplies
a zero-meter diameter if `well_diam` is absent. By default, a local UTM CRS is
selected from the combined spatial extent.

```{r prepare-spatial-inputs}
analysis_crs <- isw:::.select_analysis_crs(
  pumping_wells,
  stream_reaches
)

spatial_inputs <- isw:::.prepare_spatial_inputs(
  pumping_wells,
  stream_reaches,
  analysis_crs = analysis_crs
)

spatial_inputs$analysis_crs$Name
spatial_inputs$pumping_wells$well_diam
```

Use `analysis_crs` in the top-level call when a particular projected CRS is
required. This can be important when inputs span UTM zones or must align with
another analysis. A geographic longitude-latitude CRS is not accepted because
the model requires linear distances.

#### 2. Discretize each input reach

The function divides every reach into approximately equal segments no longer
than `reach_spacing`. Each segment retains its parent `reach_id`, receives a
generated `reach_segment_id`, and stores its actual `represented_length`.
`model_point` is the along-line midpoint used later as the equivalent
stream-injection location.

This creates three distinct spatial units in the ADF workflow:

| Spatial object | Role in the analysis |
|:---------------|:---------------------|
| **Reach** | A user-supplied source geometry identified by `reach_id`. Reaches can be used to group or summarize results, but analytical responses are not calculated directly at this level. |
| **Reach segment** | The primary ADF computational unit, created from each reach according to `reach_spacing`. Each pump--segment pair receives an `apportionment_fraction` and `pump_to_reach_distance`. Depletion is returned by segment; for drawdown, `model_point` becomes an equivalent injection well and `represented_length` becomes its effective diameter. |
| **Sample point** | A temporary weighting location created within a segment according to `sample_spacing`. Its distance and `sampled_length` determine a web or web-squared weight. These weights are summed to the segment before time-dependent depletion and drawdown are calculated. |

```{r discretize-stream-reaches}
reach_segments <- isw:::.discretize_stream_reaches(
  spatial_inputs$stream_reaches,
  reach_spacing = set_units(400, "m")
)

reach_segments[c(
  "reach_id",
  "reach_segment_id",
  "represented_length"
)]
```

```{r plot-reach-segments}
ggplot() +
  geom_sf(data = spatial_inputs$stream_reaches, color = "grey75", linewidth = 3) +
  geom_sf(data = reach_segments, aes(color = reach_segment_id), linewidth = 1.5) +
  geom_sf(
    data = reach_segments,
    aes(geometry = model_point),
    shape = 21,
    fill = "white",
    color = "black",
    size = 2.5
  ) +
  geom_sf(data = spatial_inputs$pumping_wells, color = "firebrick", size = 3) +
  labs(color = "Reach segment") +
  theme_minimal()
```

Specifying a smaller `reach_spacing` creates more model segments. Doing so
would resolve the location and timing of depletion more finely, but it
increases output size, the number of analytical responses, and computational
burden. Segment lengths can be shorter than the requested maximum because each
source line component is divided into equal parts.

#### 3. Generate points for spatial weighting

These points approximate the continuous distance-based weighting along each
reach segment. `generate_segment_sample_points()` divides each segment into
equal-length intervals, places one point at each interval center, and records
the length represented by that point as `sampled_length`. Point weights are
later summed by reach segment to calculate static apportionment fractions.
These points are distinct from the segment's single `model_point`, which is
used in the drawdown calculation.

```{r generate-segment-sample-points}
sample_points <- generate_segment_sample_points(
  reach_segments,
  sample_spacing = set_units(100, "m")
)

sample_points[c(
  "reach_id",
  "reach_segment_id",
  "sample_point_id",
  "sampled_length"
)]
```

```{r plot-sample-points}
ggplot() +
  geom_sf(data = reach_segments, color = "grey65", linewidth = 1.5) +
  geom_sf(data = sample_points, color = "royalblue", size = 1.5) +
  geom_sf(
    data = reach_segments,
    aes(geometry = model_point),
    shape = 21,
    fill = "white",
    color = "black",
    size = 3
  ) +
  geom_sf(data = spatial_inputs$pumping_wells, color = "firebrick", size = 3) +
  theme_minimal()
```

A smaller `sample_spacing` follows bends in the stream more closely when
approximating the distance-weighted integral. It improves the static spatial
approximation without increasing the number of reach segments returned by the
model. Every segment receives at least one sample point.

#### 4. Weight points and aggregate them to segments

For sample point \(j\), the unnormalized weight is

\[
w_j = \frac{L_j}{d_j^p},
\]

where \(L_j\) is `sampled_length`, \(d_j\) is the pump-to-point distance, and
\(p\) is 1 for `method = "web"` or 2 for `method = "web_squared"`. Weights are
normalized separately for each pump and summed by `reach_segment_id`.

`web_squared` concentrates weight more strongly near a pump; `web` produces a
broader allocation. The optional `maximum_distance` excludes sample points
beyond a specified distance. If it excludes every point for a pump, the
function stops instead of silently returning an invalid allocation.

`pump_to_reach_distance` is calculated separately as the shortest distance
from the pump to the segment line. It does not depend on `sample_spacing` and
is used as the distance input to the Glover response in
`get_apportioned_stream_depletion()`. It therefore controls the timing and
magnitude of the segment's time-dependent depletion response. It is not used
to calculate the static `apportionment_fraction`, which instead uses distances
from the pump to all sample points.

The following code reproduces these calculations for `pump_1`. It exposes the
point-level weights that `get_stream_reach_apportionment()` normally aggregates.
The example has no zero-distance points and does not apply `maximum_distance`;
`get_stream_reach_apportionment()` handles both cases before normalizing the
weights.

```{r inspect-point-weights}
pump_1 <- spatial_inputs$pumping_wells[1, ]
point_distances <- st_distance(pump_1, sample_points)[1, ]
segment_distances <- st_distance(pump_1, reach_segments)[1, ]
segment_distance_table <- tibble(
  reach_segment_id = reach_segments$reach_segment_id,
  pump_to_reach_distance = segment_distances
)

point_weight_table <- sample_points |>
  st_drop_geometry() |>
  mutate(
    pump_to_point_distance = point_distances,
    raw_weight = as.numeric(sampled_length) /
      as.numeric(pump_to_point_distance)^2,
    point_fraction = raw_weight / sum(raw_weight)
  )

segment_weight_table <- point_weight_table |>
  group_by(reach_id, reach_segment_id) |>
  summarize(
    apportionment_fraction = sum(point_fraction),
    .groups = "drop"
  ) |>
  left_join(segment_distance_table, by = "reach_segment_id")

head(point_weight_table)
segment_weight_table
```

Because each pump is apportioned independently, its fractions sum to one.

```{r check-apportionment-balance}
stream_apportionment |>
  st_drop_geometry() |>
  group_by(pump_id) |>
  summarize(
    total_apportionment_fraction = sum(apportionment_fraction),
    .groups = "drop"
  )
```

```{r plot-stream-apportionment}
ggplot(stream_apportionment) +
  geom_sf(aes(color = apportionment_fraction), linewidth = 2) +
  geom_sf(data = pumping_wells, color = "firebrick", size = 2.5) +
  facet_wrap(~pump_id) +
  labs(color = "Apportionment\nfraction") +
  theme_minimal()
```

## Calculate apportioned stream depletion

The next ADF modeling step combines the static spatial allocation with the
time-dependent Glover response to the pumping schedule.

```{r calculate-apportioned-stream-depletion}
stream_depletion <- get_apportioned_stream_depletion(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  evaluation_times
)

stream_depletion
```

### What happens inside `get_apportioned_stream_depletion()`

#### 1. Convert scheduled rates to change events

Analytical groundwater equations used in the package are linear and follow the
principle of superposition. This allows an intermittent pumping schedule to be
represented as a series of changes in pumping rate summed together. The first
scheduled rate is treated as a change from zero pumping, and each later rate is
compared with the preceding rate. Events with no change in pumping rate are
omitted.

```{r pumping-rate-changes}
pumping_rate_changes <- isw:::.get_pumping_rate_changes(
  pumping_schedules,
  pumping_wells
)

pumping_rate_changes
```

#### 2. Pair events with later evaluations

Every nonzero rate change is paired with evaluation times strictly later than
the event, and elapsed time is calculated in days. A change beginning exactly
at an evaluation time has not yet produced a response at that time.

```{r pumping-response-times}
pumping_response_times <- isw:::.get_pumping_response_times(
  pumping_schedules,
  pumping_wells,
  evaluation_times
)

head(pumping_response_times)
```

#### 3. Evaluate and reuse analytical fractions

The Glover stream-depletion fraction depends on the pump's aquifer properties,
its shortest distance to a reach segment, and elapsed time. The function
evaluates every unique pump--segment--elapsed-time combination once and reuses
the result for repeated event rows.

```{r depletion-fraction-lookup}
depletion_fraction_lookup <- isw:::.get_stream_depletion_fraction_lookup(
  pumping_wells,
  pumping_response_times,
  stream_apportionment
)

head(depletion_fraction_lookup)
```

For event \(e\), pump \(p\), and segment \(s\), the contribution is

\[
\Delta q_{p,s,e} = \Delta Q_{p,e} F_{p,s}(t_e) A_{p,s},
\]

where \(\Delta Q\) is the signed pumping-rate change, \(F\) is the analytical
depletion fraction, and \(A\) is the static apportionment fraction. Earlier
event contributions are summed by superposition. The output retains separate
rows for every pump, evaluation time, and reach segment, including explicit
zeroes where no earlier event contributes.

The principal user choice at this stage is `evaluation_times`: a finer grid
returns a smoother time series but creates more output rows and response
combinations. It does not change the pumping schedule or static apportionment.
The pumping schedule is plotted above the depletion response over the
same annual horizon. The final zero-rate schedule row is extended through the
last evaluation time to make the shutdown period explicit.

```{r summarize-stream-depletion, fig.height=9}
total_stream_depletion <- stream_depletion |>
  group_by(pump_id, evaluation_time) |>
  summarize(
    stream_depletion_rate = sum(stream_depletion_rate),
    .groups = "drop"
  )

annual_pumping_schedule <- bind_rows(
  pumping_schedules,
  pumping_schedules[nrow(pumping_schedules), ] |>
    mutate(t = max(evaluation_times))
) |>
  distinct(t, .keep_all = TRUE) |>
  tidyr::pivot_longer(
    cols = -t,
    names_to = "pump_id",
    values_to = "pumping_rate"
  )

pumping_schedule_plot <- ggplot(
  annual_pumping_schedule,
  aes(t, pumping_rate, color = pump_id)
) +
  geom_step(linewidth = 1) +
  labs(
    x = "Date",
    y = "Pumping rate",
    color = "Pumping well"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

stream_depletion_plot <- ggplot(
  total_stream_depletion,
  aes(evaluation_time, stream_depletion_rate, color = pump_id)
) +
  geom_line(linewidth = 1) +
  labs(
    x = "Evaluation time",
    y = "Stream-depletion rate",
    color = "Pumping well"
  ) +
  theme_minimal()

patchwork::wrap_plots(
  pumping_schedule_plot,
  stream_depletion_plot,
  ncol = 1
)
```

## Generate the stream-injection schedule for ADF drawdown

In order to estimate aquifer drawdown, we represent stream depletion as
injection wells. Each reach segment becomes an equivalent injection well whose
schedule is derived from the depletion apportioned to that segment.
`get_stream_injection_schedule()` returns one row per pump, segment, and time
interval. Each row specifies the constant injection rate applied at the
segment's `model_point` during that interval.

```{r construct-stream-injection-schedule}
stream_injection_schedule <- get_stream_injection_schedule(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  evaluation_times
)

head(stream_injection_schedule)
```

### What happens inside `get_stream_injection_schedule()`

#### 1. Construct the injection time grid

The internal grid combines every requested evaluation time with all
pumping-schedule times through the final evaluation. Each time starts a new
injection interval, except the final time, which closes the last interval.
Including pumping-schedule times ensures that a change in pumping begins a new
interval even when results are requested less frequently. If the first
evaluation occurs after pumping begins, the first pumping-schedule time is also
included and depletion is assumed to be zero at that initial time.

`.get_stream_injection_times()` returns `injection_grid`, a list containing the
requested evaluation times, the combined injection times, and those injection
times converted to elapsed days. The combined injection times define the
intervals used below.

```{r injection-time-grid}
injection_grid <- isw:::.get_stream_injection_times(
  pumping_schedules,
  evaluation_times,
  injection_times = NULL
)

tibble(
  component = names(injection_grid),
  number_of_times = lengths(injection_grid)
)
head(injection_grid$injection_times)
tail(injection_grid$injection_times)
```

#### 2. Approximate continuously changing depletion by intervals

Depletion is evaluated at the start and end of each interval. The negative
arithmetic mean of those two depletion rates becomes the constant injection
rate for that interval. The negative sign represents water entering the
aquifer from the stream, opposite to the sign used for pumping.

```{r internal-injection-schedule}
depletion_at_injection_times <- get_apportioned_stream_depletion(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  injection_grid$injection_times
)

head(depletion_at_injection_times)

internal_injection_schedule <-
  isw:::.get_interval_average_injection_schedule(
    depletion_at_injection_times,
    pumping_schedules
  )

injection_rate_changes <- isw:::.get_injection_rate_changes(
  internal_injection_schedule
)

head(internal_injection_schedule)
head(injection_rate_changes)
```

`depletion_at_injection_times` contains the segment-level depletion rates at
the start and end of each interval.
`.get_interval_average_injection_schedule()` converts them to the interval
rates returned by `get_stream_injection_schedule()`.
`.get_injection_rate_changes()` then converts successive interval rates to
change events for the superposition calculation in
`get_apportioned_aquifer_drawdown()`.

Optional `injection_times` add times at which new intervals start, creating a
finer piecewise-constant approximation. They do not change `evaluation_times`.
Extra times after the final evaluation time are ignored.

Most users can let `get_apportioned_aquifer_drawdown()` create the schedule
internally. Creating it explicitly is useful when it needs to be inspected,
reused, or held fixed while other drawdown inputs are varied.

## Estimate aquifer drawdown and stream recovery

The final ADF modeling step calculates direct pumping drawdown, recovery caused
by the equivalent stream injections, and their net effect.

```{r calculate-apportioned-aquifer-drawdown}
water_level_changes <- get_apportioned_aquifer_drawdown(
  pumping_wells,
  pumping_schedules,
  observation_wells,
  stream_apportionment,
  evaluation_times,
  stream_injection_schedule = stream_injection_schedule
)

water_level_changes
```

### What happens inside `get_apportioned_aquifer_drawdown()`

#### 1. Create physical and equivalent well geometries

Pumping and observation wells are transformed to the projected CRS stored by
`stream_apportionment`. Direct pumping distances are measured from each
physical pumping well to each observation well. Each reach segment's
`model_point` becomes an equivalent injection well, and its distance to every
observation well is calculated.

The physical response uses the user-supplied pumping-well diameter. An
injection point instead uses its segment's `represented_length` as an effective
diameter. This holds the point-well response constant inside half a segment
length and prevents a singular response at the stream.

These intermediate geometries and distance matrices can be constructed and
inspected directly:

```{r prepare-drawdown-geometries}
drawdown_crs <- st_crs(stream_apportionment)
prepared_pumping_wells <- pumping_wells |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_transform(drawdown_crs)
prepared_observation_wells <- observation_wells |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_transform(drawdown_crs)

injection_points <- st_sf(
  pump_id = stream_apportionment$pump_id,
  reach_segment_id = stream_apportionment$reach_segment_id,
  well_diam = stream_apportionment$represented_length,
  geometry = st_transform(stream_apportionment$model_point, drawdown_crs)
)

pump_to_observation_distances <- st_distance(
  prepared_pumping_wells,
  prepared_observation_wells
)
injection_to_observation_distances <- st_distance(
  injection_points,
  prepared_observation_wells
)

pump_to_observation_distances
head(injection_points)
head(injection_to_observation_distances)
```

#### 2. Convert pumping and injection schedules to events

Physical pumping uses the rate-change events demonstrated earlier. The stream
schedule is also converted from interval rates to injection-rate changes so
both responses can use the same superposition principle. Because this example
supplies `stream_injection_schedule`, the function validates and reuses it
rather than calculating depletion again.

The two helper functions return different forms of event data.
`.get_pumping_response_times()` converts the pumping schedule to rate changes,
pairs each change with every later evaluation time, and calculates the elapsed
response time for each pair. `.get_injection_rate_changes()` only converts the
interval injection schedule to changes in injection rate. Injection events are
kept in this compact form because there is one event series for every pump and
reach segment. During the drawdown calculation, each evaluation time is paired
with its earlier injection events and their elapsed times are calculated then.

If `stream_injection_schedule = NULL`, the function constructs one internally.
In that case, optional `injection_times` may refine its time grid. The two
arguments cannot be supplied together because an explicit schedule already
defines its own intervals.

The physical and injection event tables are available for inspection as
follows. Their different columns reflect when the pairing with evaluation
times occurs.

```{r inspect-drawdown-events}
physical_pumping_events <- isw:::.get_pumping_response_times(
  pumping_schedules,
  pumping_wells,
  evaluation_times
)
stream_injection_events <- isw:::.get_injection_rate_changes(
  stream_injection_schedule
)

head(physical_pumping_events)
head(stream_injection_events)
```

#### 3. Evaluate Theis responses and combine them

For every pump, observation well, and evaluation time, the function sums the
Theis responses from all earlier physical pumping events. It separately sums
responses from all earlier injection events assigned to that pump. Both use
the aquifer properties `K`, `D`, and `V` of the originating pumping well.

`pumping_drawdown` and `stream_recovery` are reported as positive magnitudes.
`water_level_change` uses a signed convention in which falling water levels are
negative and rising water levels are positive:

\[
\text{water_level_change} =
-\text{pumping_drawdown} + \text{stream_recovery}.
\]

Results remain separated by `pump_id` and `observation_id`, allowing users to
inspect individual contributions or sum across pumps. Observation locations,
evaluation times, injection-grid resolution, pumping-well diameter, and
aquifer properties are therefore the main choices controlling this step.

```{r plot-apportioned-aquifer-drawdown}
water_level_changes |>
  mutate(
    pumping_water_level_change = -pumping_drawdown
  ) |>
  tidyr::pivot_longer(
    cols = c(
      pumping_water_level_change,
      stream_recovery,
      water_level_change
    ),
    names_to = "component",
    values_to = "water_level"
  ) |>
  ggplot(aes(
    evaluation_time,
    water_level,
    color = component,
    linetype = component
  )) +
  geom_line(linewidth = 0.9) +
  scale_linetype_manual(values = c(
    pumping_water_level_change = "dashed",
    stream_recovery = "dashed",
    water_level_change = "solid"
  )) +
  facet_grid(observation_id ~ pump_id) +
  labs(
    x = "Evaluation time",
    y = "Water-level change",
    color = "Component",
    linetype = "Component"
  ) +
  theme_minimal()
```
