---
title: "Stream Depletion Apportionment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stream Depletion Apportionment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages}
library(isw)
library(units)
library(tibble)
library(dplyr)
library(ggplot2)
library(sf)
```

## Introduction

This vignette develops a stream-depletion apportionment example from the
package's spatial and pumping-schedule inputs. We begin with the inputs and
functions that are currently available. At this stage, depletion is calculated
using the distance to the closest stream reach. Later sections will distribute
that depletion among the discretized stream reaches.

## Define the pumping well

`pumping_wells` is an `sf` point object. Each pumping well has a unique
`pump_id` and the aquifer properties associated with that well. Aquifer
properties retain their physical units.

```{r pumping-wells}
pumping_wells <- st_as_sf(
  tibble(
    pump_id = "pump_1",
    x = 499800,
    y = 4980050,
    K = set_units(1e-5, "m/s"),
    D = set_units(50, "m"),
    V = 0.1
  ),
  coords = c("x", "y"),
  crs = 32615
)

pumping_wells
```

## Define the pumping schedule and evaluation times

`pumping_schedules` uses wide format. Its first column contains the schedule
times, and every remaining column is named using a `pump_id` from
`pumping_wells`. The pumping rate begins at its corresponding schedule time.

This first example uses one constant pumping event of 500 cubic meters per day.
We evaluate its response at the end of each of the following 100 days.

```{r pumping-schedule}
pumping_schedules <- tibble(
  t = as.Date("2025-01-01"),
  pump_1 = set_units(500, "m^3/day")
)

evaluation_times <- seq.Date(
  from = pumping_schedules$t + 1,
  by = "day",
  length.out = 100
)

pumping_schedules
head(evaluation_times)
```

## Define the stream network

`stream_reaches` is an `sf` line object with one row and one unique `reach_id`
for each input stream reach. Here, two upstream reaches meet at a confluence
and flow into a downstream reach.

```{r stream-reaches}
stream_reaches <- st_sf(
  reach_id = c("upstream_1", "upstream_2", "downstream"),
  geometry = st_sfc(
    st_linestring(
      matrix(c(500000, 4980200, 500150, 4980050), ncol = 2, byrow = TRUE)
    ),
    st_linestring(
      matrix(c(500000, 4979900, 500150, 4980050), ncol = 2, byrow = TRUE)
    ),
    st_linestring(
      matrix(c(500150, 4980050, 500350, 4979850), ncol = 2, byrow = TRUE)
    ),
    crs = 32615
  )
)

ggplot() +
  geom_sf(data = stream_reaches, linewidth = 1) +
  geom_sf(data = pumping_wells, color = "red", size = 3) +
  theme_minimal()
```

## Prepare the spatial inputs

The spatial inputs can be supplied in any defined coordinate reference system.
For distance-based calculations, they are transformed to one projected
analysis CRS. Because `well_diam` was not supplied, the prepared pumping-well
object assigns it a default value of zero meters.

The spatial-preparation function is currently internal. It will eventually be
called by the public stream-depletion interface.

```{r prepare-spatial-inputs}
spatial_inputs <- isw:::.prepare_spatial_inputs(
  pumping_wells,
  stream_reaches
)

spatial_inputs$analysis_crs$Name
spatial_inputs$pumping_wells$well_diam
```

## Discretize the stream network

Each input reach is divided into reach segments that do not exceed the requested
spacing. The active geometry remains a line segment, while `model_point`
contains the point at the center of each reach segment. `represented_length`
records the actual stream length represented by that point.

```{r discretize-stream-reaches}
model_reaches <- isw:::.discretize_stream_reaches(
  spatial_inputs$stream_reaches,
  reach_spacing = set_units(100, "m")
)

model_reaches[c(
  "reach_id",
  "reach_segment_id",
  "represented_length"
)]
```

```{r plot-discretized-stream-network}
ggplot() +
  geom_sf(
    data = spatial_inputs$stream_reaches,
    color = "grey75",
    linewidth = 3
  ) +
  geom_sf(
    data = model_reaches,
    aes(color = reach_segment_id),
    linewidth = 1.5
  ) +
  geom_sf(
    data = model_reaches,
    aes(geometry = model_point),
    color = "black",
    fill = "white",
    shape = 21,
    size = 2.5
  ) +
  geom_sf(
    data = spatial_inputs$pumping_wells,
    color = "red",
    size = 3
  ) +
  labs(color = "Reach segment") +
  theme_minimal()
```

## Prepare the pumping response times

The pumping schedule is converted into changes in pumping rate. Each nonzero
change is then paired with the evaluation times at which it can affect stream
depletion. `elapsed_time` is the time passed to the analytical response
function.

```{r prepare-pumping-response-times}
pumping_response_times <- isw:::.get_pumping_response_times(
  pumping_schedules,
  spatial_inputs$pumping_wells,
  evaluation_times
)

pumping_response_times
```

## Calculate depletion from the closest stream

Before adding apportionment, we can reproduce the conventional analytical
calculation by treating the closest reach as the single infinite stream. The
distance is calculated from the pumping well to the original stream geometry.

```{r closest-stream-distance}
pump_to_reach_distances <- st_distance(
  spatial_inputs$pumping_wells,
  spatial_inputs$stream_reaches
)

distance_to_stream <- min(pump_to_reach_distances)
distance_to_stream
```

The aquifer properties are joined to the pumping-event table using `pump_id`.
The descriptive `distance_to_stream` variable is passed to the existing Glover
kernel as its `x1` argument.

```{r calculate-stream-depletion}
df <- pumping_response_times |>
  left_join(
    st_drop_geometry(spatial_inputs$pumping_wells),
    by = "pump_id"
  ) |>
  mutate(
    distance_to_stream = distance_to_stream,
    stream_depletion_fraction = get_stream_depletion_fraction(
      x1 = distance_to_stream,
      K = K,
      D = D,
      V = V,
      t = elapsed_time
    ),
    stream_depletion_rate =
      stream_depletion_fraction * pumping_rate_change
  )

df |>
  select(
    pump_id,
    evaluation_time,
    elapsed_time,
    distance_to_stream,
    stream_depletion_fraction,
    stream_depletion_rate
  )
```

```{r plot-stream-depletion}
ggplot(df, aes(evaluation_time, stream_depletion_rate)) +
  geom_line() +
  labs(
    x = "Evaluation time",
    y = "Stream-depletion rate"
  ) +
  theme_minimal()
```

This calculation uses only the closest stream distance and does not yet
distribute depletion among the reach segments. The next step is to construct
the pump--reach-segment pairs needed for depletion apportionment.
