---
title: "Stream Depletion Apportionment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stream Depletion Apportionment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5.5,
  out.width = "100%",
  fig.align = "center"
)
```

```{r packages}
library(isw)
library(units)
library(tibble)
library(dplyr)
library(ggplot2)
library(sf)
```

## Introduction

This vignette presents the workflow for distributing pumping-induced
stream depletion across a stream network and estimating the associated
aquifer drawdown. It uses synthetic spatial inputs included with `isw` so
the example can focus on the package workflow.

For a step-by-step treatment of spatial preparation, stream discretization,
sampling, pumping events, and the analytical response calculation, see
`vignette("stream-depletion-apportionment-internals")`.

## Load the example spatial inputs

The package includes example spatial objects with two pumping wells (`example_pumping_wells`), three connected stream reaches (`example_stream_reaches`), and two observation wells (`example_observation_wells`). Pumping wells carry aquifer parameters (hydraulic conductivity,
aquifer thickness, and drainable porosity) used by the analytical response
functions.

```{r spatial-inputs}
pumping_wells <- example_pumping_wells
stream_reaches <- example_stream_reaches
observation_wells <- example_observation_wells

ggplot() +
  geom_sf(data = stream_reaches, color = "steelblue", linewidth = 1) +
  geom_sf(data = pumping_wells, color = "firebrick", size = 3) +
  geom_sf(data = observation_wells, color = "black", size = 3) +
  geom_sf_text(data = pumping_wells, aes(label = pump_id), nudge_y = 25) +
  geom_sf_text(
    data = observation_wells,
    aes(label = observation_id),
    nudge_y = -25
  ) +
  theme_minimal()
```

The `pumping_wells` object contains a `pump_id` column and aquifer parameters. Note that the `tibble` includes units for `K` and `D`. An optional column `well_diam` can be specified, such that aquifer drawdown inside the pumping well radius (`well_diam/2`) is equivalent to the estimated aquifer drawdown at edge of the well.

```{r}
pumping_wells
```

Note that the `stream_reaches` object contains a `reach_id` column.

```{r}
stream_reaches
```

## Define pumping schedules and evaluation times

Pumping timeseries are supplied to the model using pumping schedules. This should be a `tibble`  provided in wide format. The first column, `t`, represents the start of a new pumping period for the row. All other columns should have names that match the `pump_id` values in the `pumping_wells` object. A schedule row
therefore describes the rate applied by every modeled pump from that time
until the next schedule change (i.e., the next row).

Evaluation times are separate from schedule times. They determine when model
results are returned, without requiring pumping rates to change at every
evaluation.

```{r pumping-schedule}
pumping_schedules <- tibble(
  t = as.Date(c("2025-01-01", "2025-02-01", "2025-03-01", "2025-04-01")),
  pump_1 = set_units(c(500, 500, 300, 0), "m^3/day"),
  pump_2 = set_units(c(0, 250, 250, 0), "m^3/day")
)

evaluation_times <- seq.Date(
  from = pumping_schedules$t[1] + 1,
  by = "day",
  length.out = 120
)

pumping_schedules
```

## Apportioning depletion among stream reaches

The ADF approach involve apportioning stream depletion to each reach of the river.
`get_stream_reach_apportionment()` first transforms the spatial inputs to a
common projected analysis CRS. It divides each reach into modeled segments no
longer than `reach_spacing` and represents the detailed geometry of each
segment with points separated by at most `sample_spacing`.

The web-squared method weights the sampled stream lengths by inverse squared
distance from each pump. These weights are summed by reach segment and
normalized independently for every pumping well. The result also stores the
shortest pump-to-segment distance used by the time-dependent analytical
stream-depletion calculation.

```{r calculate-stream-apportionment}
stream_apportionment <- get_stream_reach_apportionment(
  pumping_wells,
  stream_reaches,
  reach_spacing = set_units(100, "m"),
  sample_spacing = set_units(25, "m"),
  method = "web_squared"
)

stream_apportionment[c(
  "pump_id",
  "reach_id",
  "reach_segment_id",
  "pump_to_reach_distance",
  "apportionment_fraction"
)]
```

Because each pump is apportioned separately, its fractions sum to one across
the complete network.

```{r check-apportionment-balance}
stream_apportionment |>
  st_drop_geometry() |>
  group_by(pump_id) |>
  summarize(
    total_apportionment_fraction = sum(apportionment_fraction),
    .groups = "drop"
  )
```

```{r plot-stream-apportionment}
ggplot(stream_apportionment) +
  geom_sf(aes(color = apportionment_fraction), linewidth = 2) +
  geom_sf(data = pumping_wells, color = "firebrick", size = 2.5) +
  facet_wrap(~pump_id) +
  labs(color = "Apportionment\nfraction") +
  theme_minimal()
```

## Calculate apportioned stream depletion

`get_apportioned_stream_depletion()` converts each pumping schedule into
nonzero changes in pumping rate and pairs those changes with later evaluation
times. It evaluates the Glover response only for unique combinations of pump,
reach segment, and elapsed time. Each event response is multiplied by the
static reach apportionment fraction, and responses from all earlier pumping
events are combined by superposition.

The returned values remain separated by pump and reach segment. This preserves
the provenance of each contribution before totals are calculated.

```{r calculate-apportioned-stream-depletion}
stream_depletion <- get_apportioned_stream_depletion(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  evaluation_times
)

stream_depletion
```

Total depletion can be summarized either by pump or across the complete stream
network. Here it is summed across reach segments while retaining each pump.

```{r summarize-stream-depletion}
total_stream_depletion <- stream_depletion |>
  group_by(pump_id, evaluation_time) |>
  summarize(
    stream_depletion_rate = sum(stream_depletion_rate),
    .groups = "drop"
  )

ggplot(
  total_stream_depletion,
  aes(evaluation_time, stream_depletion_rate, color = pump_id)
) +
  geom_line() +
  labs(
    x = "Evaluation time",
    y = "Stream-depletion rate",
    color = "Pumping well"
  ) +
  theme_minimal()
```

## Inspect the stream-injection schedule

Aquifer recovery from stream depletion is represented by injection at each
reach segment's model point. `get_stream_injection_schedule()` evaluates
depletion at the boundaries of each pumping interval and uses the negative
arithmetic mean of the two endpoint rates as a piecewise-constant injection
rate. Optional injection times can add temporal resolution without changing
the requested output times.

Most users can let the aquifer-drawdown function construct this schedule
internally. It is exposed so it can be inspected, refined, or reused.

For the drawdown calculation, each point's effective injection-well diameter
is its segment's actual `represented_length`. This prevents a point-source
singularity at the stream.

```{r construct-stream-injection-schedule}
stream_injection_schedule <- get_stream_injection_schedule(
  pumping_wells,
  pumping_schedules,
  stream_apportionment,
  evaluation_times
)

head(stream_injection_schedule)
```

## Estimate aquifer drawdown and stream recovery

`get_apportioned_aquifer_drawdown()` superimposes two sets of Theis responses.
Physical pumping wells produce positive drawdown at every observation well,
while the segment injection wells produce positive stream recovery.
`water_level_change` is negative when pumping lowers the water level and
positive when stream recovery raises it.

The calculation remains separated by originating pump, observation well, and
evaluation time. Supplying the schedule calculated above avoids repeating the
stream-depletion calculation and uses exactly the inspected injection grid.

```{r calculate-apportioned-aquifer-drawdown}
water_level_changes <- get_apportioned_aquifer_drawdown(
  pumping_wells,
  pumping_schedules,
  observation_wells,
  stream_apportionment,
  evaluation_times,
  stream_injection_schedule = stream_injection_schedule
)

water_level_changes
```

```{r plot-apportioned-aquifer-drawdown}
water_level_changes |>
  mutate(
    pumping_water_level_change = -pumping_drawdown
  ) |>
  tidyr::pivot_longer(
    cols = c(
      pumping_water_level_change,
      stream_recovery,
      water_level_change
    ),
    names_to = "component",
    values_to = "water_level"
  ) |>
  ggplot(aes(evaluation_time, water_level, color = component)) +
  geom_line() +
  facet_grid(observation_id ~ pump_id) +
  labs(
    x = "Evaluation time",
    y = "Water-level change",
    color = "Component"
  ) +
  theme_minimal()
```
