---
title: "Finite Line-Sink Numerical Validation"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.5,
  out.width = "90%",
  fig.align = "center"
)
```

```{r packages, message = FALSE}
library(isw)
library(units)
library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Purpose

The finite line-sink calculation uses fixed Gauss--Legendre quadrature. This
article compares that calculation with an independent adaptive numerical
integration and makes the numerical results visible for review. It is a
diagnostic article rather than a pass/fail unit test: the tables and figures
show how the approximation changes with quadrature order and observation
location.

The comparison uses the package convention that a segment rate is a total
volumetric rate. Both calculations therefore evaluate the normalized average
response over the line. The stream width supplies an effective minimum radius
of half the width.

## Independent adaptive reference

The reference function below evaluates

\[
R_{line} = \int_{-1/2}^{1/2}
-\frac{E_1\!\left(r_{eff}^2/(4\alpha t)\right)}{4\pi K D}
\,d\xi,
\]

where \(\alpha=KD/V\), \(r_{eff}=\max(r, w/2)\), and
\(r^2=(x-L\xi)^2+y^2\). It uses `stats::integrate()` rather than the
Gauss--Legendre implementation in `isw`.

```{r reference-function}
adaptive_line_response <- function(
    along_distance,
    perpendicular_distance,
    line_length,
    K,
    D,
    V,
    elapsed_time,
    stream_width) {

  x <- drop_units(set_units(along_distance, "m"))
  y <- drop_units(set_units(perpendicular_distance, "m"))
  line_length <- drop_units(set_units(line_length, "m"))
  K <- drop_units(set_units(K, "m/day"))
  D <- drop_units(set_units(D, "m"))
  elapsed_time <- drop_units(set_units(elapsed_time, "day"))
  minimum_radius <- drop_units(set_units(stream_width, "m")) / 2
  diffusivity <- K * D / V

  integrand <- function(line_fraction) {
    distance <- sqrt((x - line_length * line_fraction)^2 + y^2)
    effective_distance <- pmax(distance, minimum_radius)
    dimensionless_time <- effective_distance^2 /
      (4 * diffusivity * elapsed_time)

    -expint::expint_E1(dimensionless_time) / (4 * pi * K * D)
  }

  reference <- stats::integrate(
    integrand,
    lower = -0.5,
    upper = 0.5,
    subdivisions = 2000,
    rel.tol = 1e-10,
    stop.on.error = TRUE
  )$value

  set_units(reference, "day/m^2")
}
```

## Evaluation cases

The cases exercise the segment midpoint, an off-center point very near the
line, an endpoint, an ordinary off-line location, and the far field. The
default 1 m stream width is used throughout so its numerical consequences are
visible. The line is 100 m long and centered at the local origin.

```{r cases}
hydraulics <- list(
  line_length = set_units(100, "m"),
  K = set_units(10, "m/day"),
  D = set_units(20, "m"),
  V = 0.15,
  elapsed_time = set_units(10, "day"),
  stream_width = set_units(1, "m")
)

cases <- tribble(
  ~location,              ~along_m, ~perpendicular_m,
  "Midpoint on line",             0,              0,
  "Off-center near line",        20,            0.2,
  "Segment endpoint",            50,              2,
  "Ordinary off-line point",     20,             20,
  "Far field",                    0,            500
)

quadrature_orders <- c(8L, 16L, 32L, 64L)
cases
```

## Calculate responses and errors

The following chunk is deliberately visible so parameters, cases, and
quadrature orders can be changed and rerun during review.

```{r calculate-comparison}
reference_response <- vapply(
  seq_len(nrow(cases)),
  function(case_index) {
    as.numeric(adaptive_line_response(
      along_distance = set_units(cases$along_m[[case_index]], "m"),
      perpendicular_distance =
        set_units(cases$perpendicular_m[[case_index]], "m"),
      line_length = hydraulics$line_length,
      K = hydraulics$K,
      D = hydraulics$D,
      V = hydraulics$V,
      elapsed_time = hydraulics$elapsed_time,
      stream_width = hydraulics$stream_width
    ))
  },
  numeric(1)
)

comparison <- tidyr::crossing(
  case_index = seq_len(nrow(cases)),
  quadrature_order = quadrature_orders
) |>
  mutate(
    location = cases$location[case_index],
    along_m = cases$along_m[case_index],
    perpendicular_m = cases$perpendicular_m[case_index],
    reference_day_per_m2 = reference_response[case_index],
    quadrature_day_per_m2 = vapply(
      seq_len(n()),
      function(result_index) {
        as.numeric(isw:::.line_sink_aquifer_drawdown_ratio(
          along_distance = set_units(along_m[[result_index]], "m"),
          perpendicular_distance =
            set_units(perpendicular_m[[result_index]], "m"),
          line_length = hydraulics$line_length,
          K = hydraulics$K,
          D = hydraulics$D,
          V = hydraulics$V,
          t = hydraulics$elapsed_time,
          stream_width = hydraulics$stream_width,
          quadrature_order = quadrature_order[[result_index]]
        ))
      },
      numeric(1)
    ),
    absolute_error_day_per_m2 =
      abs(quadrature_day_per_m2 - reference_day_per_m2),
    relative_error_percent = 100 * absolute_error_day_per_m2 /
      abs(reference_day_per_m2)
  ) |>
  select(-case_index)
```

## Review the values

The full table retains the signed response values as well as absolute and
relative errors.

```{r full-results}
comparison |>
  mutate(across(
    c(
      reference_day_per_m2,
      quadrature_day_per_m2,
      absolute_error_day_per_m2,
      relative_error_percent
    ),
    ~ signif(.x, 6)
  )) |>
  knitr::kable()
```

This compact table makes convergence by quadrature order easier to compare.

```{r error-table}
comparison |>
  select(location, quadrature_order, relative_error_percent) |>
  mutate(relative_error_percent = signif(relative_error_percent, 5)) |>
  pivot_wider(
    names_from = quadrature_order,
    values_from = relative_error_percent,
    names_prefix = "order_"
  ) |>
  knitr::kable(
    caption = "Absolute relative error (%) compared with adaptive integration"
  )
```

```{r error-plot}
ggplot(
  comparison,
  aes(
    x = quadrature_order,
    y = pmax(relative_error_percent, 1e-12),
    color = location
  )
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = quadrature_orders) +
  scale_y_log10() +
  labs(
    x = "Gauss-Legendre quadrature order",
    y = "Absolute relative error (%)",
    color = "Observation location"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Focus on the default order

`isw` currently uses order 16 by default. This table isolates that order
so its approximation error can be reviewed directly before deciding whether
the default should change.

```{r default-order}
comparison |>
  filter(quadrature_order == 16L) |>
  transmute(
    location,
    along_m,
    perpendicular_m,
    reference_day_per_m2 = signif(reference_day_per_m2, 7),
    quadrature_day_per_m2 = signif(quadrature_day_per_m2, 7),
    relative_error_percent = signif(relative_error_percent, 5)
  ) |>
  knitr::kable()
```

## Short-line convergence to a point well

A finite line carrying a fixed total volumetric rate should approach a point
well at the line midpoint as the line length approaches zero. This comparison
is also sensitive to an incorrect linear-rate convention: multiplying by line
length, rather than using total segment rate, would prevent convergence to the
point-well response shown here.

The point-well reference uses `well_diam = 1 m`, matching the line calculation's
`stream_width = 1 m`. The off-line midpoint case isolates perpendicular
distance, the off-center case includes both coordinate directions, and the
on-line case exercises the common minimum-radius treatment.

```{r short-line-cases}
short_line_cases <- tribble(
  ~location,              ~along_m, ~perpendicular_m,
  "Midpoint off line",             0,             20,
  "Off-center off line",          20,             20,
  "Midpoint on line",              0,              0
)

line_lengths_m <- c(100, 30, 10, 3, 1, 0.3, 0.1)
```

```{r calculate-short-line-convergence}
short_line_convergence <- tidyr::crossing(
  case_index = seq_len(nrow(short_line_cases)),
  line_length_m = line_lengths_m
) |>
  mutate(
    location = short_line_cases$location[case_index],
    along_m = short_line_cases$along_m[case_index],
    perpendicular_m = short_line_cases$perpendicular_m[case_index],
    point_distance_m = sqrt(along_m^2 + perpendicular_m^2),
    point_response_day_per_m2 = vapply(
      seq_len(n()),
      function(result_index) {
        as.numeric(isw:::.theis_aquifer_drawdown_ratio(
          distance = set_units(point_distance_m[[result_index]], "m"),
          K = hydraulics$K,
          D = hydraulics$D,
          V = hydraulics$V,
          t = hydraulics$elapsed_time,
          well_diam = hydraulics$stream_width
        ))
      },
      numeric(1)
    ),
    line_response_day_per_m2 = vapply(
      seq_len(n()),
      function(result_index) {
        as.numeric(isw:::.line_sink_aquifer_drawdown_ratio(
          along_distance = set_units(along_m[[result_index]], "m"),
          perpendicular_distance =
            set_units(perpendicular_m[[result_index]], "m"),
          line_length = set_units(line_length_m[[result_index]], "m"),
          K = hydraulics$K,
          D = hydraulics$D,
          V = hydraulics$V,
          t = hydraulics$elapsed_time,
          stream_width = hydraulics$stream_width,
          quadrature_order = 16L
        ))
      },
      numeric(1)
    ),
    relative_difference_percent = 100 *
      abs(line_response_day_per_m2 - point_response_day_per_m2) /
      abs(point_response_day_per_m2)
  ) |>
  select(-case_index)
```

```{r short-line-table}
short_line_convergence |>
  transmute(
    location,
    line_length_m,
    point_response_day_per_m2 = signif(point_response_day_per_m2, 7),
    line_response_day_per_m2 = signif(line_response_day_per_m2, 7),
    relative_difference_percent = signif(relative_difference_percent, 5)
  ) |>
  knitr::kable()
```

```{r short-line-plot}
ggplot(
  short_line_convergence,
  aes(
    x = line_length_m,
    y = pmax(relative_difference_percent, 1e-12),
    color = location
  )
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_x_log10(breaks = rev(line_lengths_m)) +
  scale_y_log10() +
  labs(
    x = "Line length (m)",
    y = "Difference from point-well response (%)",
    color = "Observation location"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

For each location, the relative difference should decrease toward zero as the
line becomes short relative to the observation distance or the shared minimum
radius.

These results should be interpreted in the context of the model's intended
accuracy and computational cost. In particular, the on-line and near-line
cases are intentionally demanding because a fixed quadrature rule may not
place a source point inside the narrow width-regularized region.
