---
title: "Finite Line-Sink Performance and Memory Scaling"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.5,
  out.width = "90%",
  fig.align = "center"
)
```

```{r packages, message = FALSE}
library(isw)
library(units)
library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
```

## Purpose

This article measures how the finite line-sink implementation scales as the
number of stream segments increases. It focuses on the core operations used by
the constant-head method:

1. Expand line segments into Gauss--Legendre source points.
2. Calculate distances from all collocation targets to all source points.
3. Evaluate and aggregate the finite-line response matrix.
4. Factor and solve the dense constant-head matrix.

The code and results are visible so the segment counts, quadrature order, and
model parameters can be changed and rerun. Elapsed times are specific to the
computer running the article; memory dimensions and asymptotic relationships
are platform-independent.

## Expected scaling

Let \(S\) be the number of stream segments, \(M\) the number of observation
targets, \(q\) the quadrature order, and \(e\) the average number of straight
edges per segment.

For ordinary observation-well calculations, the stored distance matrix has
\(M S q e\) entries. For the constant-head calculation, every segment midpoint
is a target, so \(M=S\) and the distance matrix has \(S^2 q e\) entries. The
aggregated response matrix has \(S^2\) entries, and its dense QR factorization
requires approximately cubic work in \(S\).

The table below shows the raw numeric storage implied by these dimensions for
single-edge segments at the default quadrature order. Actual R objects are
larger because they also carry units, geometry, dimensions, and other metadata.

```{r theoretical-memory}
quadrature_order <- 16L
theoretical_segment_counts <- c(100L, 250L, 500L, 1000L, 2000L)

theoretical_memory <- tibble(
  segments = theoretical_segment_counts,
  distance_values = quadrature_order * segments^2,
  distance_matrix_mib = 8 * distance_values / 1024^2,
  response_matrix_mib = 8 * segments^2 / 1024^2
) |>
  mutate(across(
    c(distance_matrix_mib, response_matrix_mib),
    ~ round(.x, 2)
  ))

knitr::kable(theoretical_memory)
```

## Benchmark setup

The benchmark uses one straight reach divided into equal 100 m segments. Each
segment therefore has one edge and 16 quadrature points. Its midpoint is used
as the constant-head collocation target. The hydraulic parameters remain fixed
across sizes.

The benchmark isolates one response matrix and one right-hand side. A complete
schedule can reuse response matrices and QR factorizations for repeated elapsed
times, but it may also evaluate additional matrices for historical injection
pulses. Consequently, these measurements describe the core cost rather than a
universal runtime for every schedule.

### Negligible-response shortcut

The Theis response uses the exponential integral \(E_1(u)\). The implementation
does not call `expint()` when the dimensionless argument \(u\) is greater than
700; it assigns a response of zero directly. At that threshold,
\(E_1(700)\) is approximately \(1.4 \times 10^{-307}\), so the omitted response
is negligible at ordinary model precision. This also avoids numerical
underflow warnings and saves the special-function calculation for distant
source--target pairs or very short elapsed times. The shortcut is applied while
evaluating response matrices, after distances and elapsed times are known.

The live benchmark stops at 400 segments so the article remains practical to
rerun interactively. The theoretical memory table extends to 2,000 segments.
Larger measured cases can be explored by editing `segment_counts`, with runtime
and memory requirements depending on the computer.

```{r benchmark-parameters}
segment_counts <- c(25L, 50L, 100L, 200L, 400L)
segment_length_m <- 100

K <- set_units(10, "m/day")
D <- set_units(20, "m")
V <- 0.15
elapsed_time <- set_units(10, "day")
```

```{r benchmark-function}
object_mib <- function(x) {
  as.numeric(object.size(x)) / 1024^2
}

elapsed <- function(expression) {
  gc()
  unname(system.time(force(expression))[["elapsed"]])
}

benchmark_segment_count <- function(number_of_segments) {
  stream_reach <- st_sf(
    reach_id = "benchmark_reach",
    geometry = st_sfc(
      st_linestring(matrix(
        c(
          0, 0,
          number_of_segments * segment_length_m, 0
        ),
        ncol = 2,
        byrow = TRUE
      )),
      crs = 32615
    )
  )
  stream_segments <- prep_stream_segments(
    stream_reach,
    reach_spacing = set_units(segment_length_m, "m")
  )
  stopifnot(nrow(stream_segments) == number_of_segments)
  collocation_targets <- st_sf(
    reach_segment_id = stream_segments$reach_segment_id,
    geometry = stream_segments$model_point
  )

  prepare_seconds <- elapsed({
    line_elements <- isw:::.prepare_line_elements(
      stream_segments,
      quadrature_order = quadrature_order
    )
  })
  distance_seconds <- elapsed({
    response_operator <- isw:::.prepare_line_response_operator(
      collocation_targets,
      line_elements
    )
  })
  response_seconds <- elapsed({
    response_matrix <- isw:::.evaluate_line_response_operator(
      response_operator,
      K = K,
      D = D,
      V = V,
      elapsed_time = elapsed_time
    )
  })
  factor_and_solve_seconds <- elapsed({
    response_qr <- qr(response_matrix)
    condition_number <- kappa(response_qr, exact = FALSE)
    solved_rates <- qr.coef(
      response_qr,
      matrix(1, nrow = number_of_segments, ncol = 1)
    )
  })

  stopifnot(
    all(is.finite(response_matrix)),
    all(is.finite(solved_rates)),
    is.finite(condition_number)
  )

  tibble(
    segments = number_of_segments,
    quadrature_points = nrow(line_elements$quadrature_points),
    distance_values = length(response_operator$distances),
    prepare_seconds,
    distance_seconds,
    response_seconds,
    factor_and_solve_seconds,
    line_elements_mib = object_mib(line_elements),
    response_operator_mib = object_mib(response_operator),
    response_matrix_mib = object_mib(response_matrix),
    qr_factorization_mib = object_mib(response_qr),
    matrix_condition_number = condition_number
  )
}
```

## Run the benchmark

Each size is run once because the larger operations are deterministic and the
purpose is to expose approximate scaling, not to compare small timing
differences between competing implementations. Rerunning the article provides
a fresh measurement on the current machine.

```{r run-benchmark}
benchmark_results <- bind_rows(lapply(
  segment_counts,
  benchmark_segment_count
))
```

## Timing results

```{r timing-table}
benchmark_results |>
  select(
    segments,
    quadrature_points,
    prepare_seconds,
    distance_seconds,
    response_seconds,
    factor_and_solve_seconds
  ) |>
  mutate(across(ends_with("seconds"), ~ signif(.x, 4))) |>
  knitr::kable()
```

```{r timing-plot}
timing_long <- benchmark_results |>
  select(segments, ends_with("seconds")) |>
  pivot_longer(
    -segments,
    names_to = "stage",
    values_to = "seconds"
  ) |>
  mutate(
    stage = recode(
      stage,
      prepare_seconds = "Prepare quadrature geometry",
      distance_seconds = "Calculate distances",
      response_seconds = "Evaluate response matrix",
      factor_and_solve_seconds = "Factor and solve matrix"
    )
  )

ggplot(
  timing_long,
  aes(
    x = segments,
    y = pmax(seconds, 1e-4),
    color = stage
  )
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_x_log10(breaks = segment_counts) +
  scale_y_log10() +
  labs(
    x = "Stream segments",
    y = "Elapsed time (seconds)",
    color = "Stage"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Memory results

The response-operator size includes the target-to-quadrature distance matrix
and the vectors needed to aggregate quadrature points back to source segments.
The objects in this table coexist during a constant-head solve, although R may
temporarily allocate additional copies while calculating responses and matrix
factorizations.

```{r memory-table}
benchmark_results |>
  select(
    segments,
    distance_values,
    line_elements_mib,
    response_operator_mib,
    response_matrix_mib,
    qr_factorization_mib
  ) |>
  mutate(across(ends_with("mib"), ~ round(.x, 3))) |>
  knitr::kable()
```

```{r memory-plot}
memory_long <- benchmark_results |>
  select(segments, ends_with("mib")) |>
  pivot_longer(
    -segments,
    names_to = "object",
    values_to = "mib"
  ) |>
  mutate(
    object = recode(
      object,
      line_elements_mib = "Prepared line elements",
      response_operator_mib = "Response operator",
      response_matrix_mib = "Response matrix",
      qr_factorization_mib = "QR factorization"
    )
  )

ggplot(
  memory_long,
  aes(x = segments, y = mib, color = object)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_x_log10(breaks = segment_counts) +
  scale_y_log10() +
  labs(
    x = "Stream segments",
    y = "Measured object size (MiB)",
    color = "Object"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Conditioning observed during the benchmark

Matrix conditioning is a property of the geometry, hydraulic parameters,
elapsed time, and segment discretization. It is shown separately because a
large condition number is a numerical-stability warning rather than a direct
measure of computational cost.

```{r condition-table}
benchmark_results |>
  transmute(
    segments,
    matrix_condition_number = signif(matrix_condition_number, 6)
  ) |>
  knitr::kable()
```

## Interpretation

The distance operator and response evaluation grow quadratically for the
constant-head geometry, while dense factorization eventually grows cubically.
The measured tables show when those costs become material on the machine used
for analysis. They can be used to decide whether the current implementation is
adequate for expected networks, whether quadrature construction should be
optimized, or whether a practical segment-count warning should be documented.
