---
title: "Comparing ADF and Constant-Head Stream Injection"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  out.width = "100%",
  fig.align = "center"
)
```

```{r packages, message = FALSE}
library(isw)
library(units)
library(tibble)
library(dplyr)
library(ggplot2)
library(sf)
```

## Purpose

`isw` supports two ways to model how a stream responds to pumping:

* `method = "adf"` estimates total stream depletion with analytical depletion
  functions, then distributes that depletion among stream segments. This is
  the default method.
* `method = "constant_head"` solves for the injection rate at each stream
  segment. The rates are chosen so that net water-level change is zero at each
  segment's model point at the end of every injection interval.

This vignette uses both methods to evaluate the same aquifer and pumping
scenario. The methods share the stream geometry, pumping schedule, temporal
grid, and observation locations so their results can be compared directly.
The scenario is deliberately simple: one pumping well beside a straight
stream in a homogeneous aquifer. This geometry also permits comparison with
the analytical image-well solution.

## Define a common model

Distance calculations require a projected CRS. This example creates its
inputs directly in UTM zone 15N. When `stream_reaches` uses a geographic CRS,
`get_stream_segments()` selects a local UTM CRS unless the user supplies
`analysis_crs`. Later functions transform pumping and observation wells to
the CRS used by the stream segments.

The full UTM coordinates are not helpful for understanding the layout, so the
example defines `origin_x` and `origin_y` at the point where the stream crosses
the pumping-well transect. Every location is written as a distance from this
origin while remaining in the projected CRS.

Each input has a distinct role in the shared workflow:

* `stream_reaches` defines the stream geometry. `get_stream_segments()` divides
  this geometry into segments that can be used with either injection method.
* `pumping_wells` identifies the pumping wells and provides their locations and
  aquifer properties. These data are needed for ADF apportionment, the stream
  depletion schedules, and the water-level calculation.
* `pumping_schedules` gives the pumping rate through time for each well.
  `get_stream_injection_schedule()` uses it to calculate the stream response,
  and `get_aquifer_water_level_change()` uses it to calculate drawdown from the
  pumping wells.

See the [detailed spatial-input discussion](stream-depletion-apportionment-internals.html#what-the-spatial-inputs-control)
for the required stream and well columns. The
[temporal-input discussion](stream-depletion-apportionment-internals.html#what-happens-to-the-temporal-inputs)
describes the pumping-schedule format. The corresponding function references
are also available with `?get_stream_segments` and
`?get_stream_injection_schedule`.

```{r model-inputs}
analysis_crs <- 32615
origin_x <- 500000
origin_y <- 4800000

stream_reaches <- st_sf(
  reach_id = "stream_1",
  geometry = st_sfc(
    st_linestring(matrix(
      c(
        origin_x, origin_y - 600,
        origin_x, origin_y + 600
      ),
      ncol = 2,
      byrow = TRUE
    )),
    crs = analysis_crs
  )
)

pumping_wells <- st_as_sf(
  tibble(
    pump_id = "pump_1",
    x = origin_x + 300,
    y = origin_y,
    K = set_units(10, "m/day"),
    D = set_units(30, "m"),
    V = 0.15,
    well_diam = set_units(0.3, "m")
  ),
  coords = c("x", "y"),
  crs = analysis_crs
)

pumping_schedules <- tibble(
  t = set_units(c(0, 120), "days"),
  pump_1 = set_units(c(400, 0), "m^3/day")
)
```

## Create neutral stream segments

`get_stream_segments()` divides the stream before either injection method is
applied. Its output contains no ADF weights or injection rates, so the same
segments can be used by both methods.

Each segment has a `model_point`. This point is the location of the segment's
discrete injection well and, for the constant-head method, the point where the
zero-change condition is enforced. By default, `well_diam` is half the length
represented by the segment. The Theis response is therefore capped within a
radius equal to one quarter of that length.

```{r stream-segments}
stream_segments <- get_stream_segments(
  stream_reaches,
  reach_spacing = set_units(100, "m")
)

stream_segments[c(
  "reach_id", "reach_segment_id", "represented_length", "well_diam"
)]
```

## Place observations on and away from the stream

Observations spaced every 25 m along the stream show the modeled water level at
the collocation points and between them. Two perpendicular transects show how
the response changes with distance from the stream. One crosses the pumping
well; the other is 300 m north.

```{r observation-wells, fig.height = 8.25}
segment_coordinates <- st_coordinates(stream_segments$model_point)
boundary_y <- seq(origin_y - 575, origin_y + 575, by = 25)
boundary_observations <- tibble(
  observation_id = paste0("boundary_", seq_along(boundary_y)),
  observation_type = "stream boundary",
  transect = NA_character_,
  distance_from_stream = set_units(0, "m"),
  distance_along_stream = set_units(boundary_y - origin_y, "m"),
  x = origin_x,
  y = boundary_y,
  is_collocation = boundary_y %in% segment_coordinates[, "Y"]
)

transect_distances <- c(0, 10, 25, 50, 100, 200, 300, 500, 800)
transect_offsets <- c(center = 0, north = 300)
transect_observations <- do.call(
  rbind,
  lapply(names(transect_offsets), function(transect_name) {
    tibble(
      observation_id = paste0(
        "transect_", transect_name, "_", transect_distances
      ),
      observation_type = "transect",
      transect = transect_name,
      distance_from_stream = set_units(transect_distances, "m"),
      distance_along_stream = set_units(
        rep(transect_offsets[[transect_name]], length(transect_distances)),
        "m"
      ),
      x = origin_x + transect_distances,
      y = origin_y + transect_offsets[[transect_name]],
      is_collocation = FALSE
    )
  })
)
observation_table <- bind_rows(
  boundary_observations,
  transect_observations
)
observation_wells <- st_as_sf(
  observation_table,
  coords = c("x", "y"),
  crs = analysis_crs
)

ggplot() +
  geom_sf(data = stream_segments, color = "grey70", linewidth = 1.4) +
  geom_sf(
    data = filter(
      observation_wells,
      observation_type == "transect"
    ),
    aes(color = "Transect observation", shape = "Transect observation"),
    size = 2
  ) +
  geom_sf(
    data = filter(
      observation_wells,
      observation_type == "stream boundary" & !is_collocation
    ),
    aes(
      color = "Other boundary observation",
      shape = "Other boundary observation"
    ),
    size = 2.4
  ) +
  geom_sf(
    data = st_sf(geometry = stream_segments$model_point),
    aes(
      color = "Injection / collocation point",
      shape = "Injection / collocation point"
    ),
    size = 5.4
  ) +
  geom_sf(
    data = pumping_wells,
    aes(color = "Pumping well", shape = "Pumping well"),
    size = 3.5
  ) +
  scale_color_manual(values = c(
    "Pumping well" = "firebrick",
    "Injection / collocation point" = "navy",
    "Other boundary observation" = "#009E73",
    "Transect observation" = "#6A3D9A"
  )) +
  scale_shape_manual(values = c(
    "Pumping well" = 8,
    "Injection / collocation point" = 18,
    "Other boundary observation" = 17,
    "Transect observation" = 1
  )) +
  coord_sf(xlim = c(origin_x - 25, origin_x + 825)) +
  labs(
    subtitle = paste(
      "The segmented stream has the same footprint as the original",
      "stream reach"
    ),
    x = NULL,
    y = NULL,
    color = NULL,
    shape = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Boundary observations that coincide with model points are shown only
as a blue injection/collocation point, rather than as two overlapping symbols.

## Construct ADF and constant-head schedules

Before constructing the schedules, we choose the output times and the time
step used for stream injection.

```{r temporal-grids}
evaluation_times <- set_units(c(30, 60, 90, 120, 180), "days")
injection_times <- set_units(seq(0, 180, by = 15), "days")
```

`evaluation_times` specifies when the model returns water-level results. The
stream-depletion schedule must extend through all of these times.

`injection_times` controls the internal time resolution of that schedule. It
adds boundaries between piecewise-constant injection intervals but does not
request more water-level output. Pumping-schedule times and evaluation times
are always included as interval boundaries, whether or not `injection_times`
is supplied.

For ADF, stream depletion changes continuously. The schedule approximates this
change with a constant rate in each interval, using the average depletion at
the interval's two endpoints. Shorter intervals improve this approximation.

For constant head, the model solves one set of injection rates for each
interval and holds those rates constant until the next interval begins. The
zero-change condition for hydraulic head at the stream is enforced at the interval end. Shorter intervals let
the rates adjust more often and enforce the condition more frequently. They
also alter the injection history carried into later intervals, so they can
change constant-head results even when `evaluation_times` stays the same. This example
uses the same 15-day increment for both methods so that differences in
the results are less likely to be caused by different time steps.

The extra
`injection_times` are optional. If `injection_times` is omitted, the package uses the
pumping-schedule and evaluation times as the interval endpoints to set constant injections.

More intervals require more analytical response evaluations for ADF and more
matrix solves for constant head. Note that the constant-head calculation reuses a
matrix factorization when the interval duration and aquifer properties are the
same. The
[temporal-input discussion](stream-depletion-apportionment-internals.html#what-happens-to-the-temporal-inputs)
provides more detail about evaluation times, and the
[schedule-grid walkthrough](stream-depletion-apportionment-internals.html#what-happens-inside-get_stream_injection_schedule)
shows how optional injection times are merged with required boundaries.

The ADF workflow first calculates pump-specific weights for the stream
segments, then passes those weights to `get_stream_injection_schedule()`. The
constant-head workflow passes the stream segments directly to the same
function. Both schedules use the interval grid defined above.

```{r injection-schedules}
stream_apportionment <- get_adf_stream_apportionment(
  pumping_wells,
  stream_segments,
  sample_spacing = set_units(25, "m"),
  method = "web_squared"
)

adf_schedule <- get_stream_injection_schedule(
  pumping_wells,
  pumping_schedules,
  stream_segments,
  evaluation_times,
  injection_times,
  method = "adf",
  stream_apportionment = stream_apportionment
)

constant_head_schedule <- get_stream_injection_schedule(
  pumping_wells,
  pumping_schedules,
  stream_segments,
  evaluation_times,
  injection_times,
  method = "constant_head"
)
```

The constant-head schedule keeps each pump's contribution separate. It also
reports two diagnostics. `boundary_residual` is the signed water-level change
remaining at a collocation point after the solve. `matrix_condition_number`
indicates how sensitive the solved rates are to small numerical or input
changes. `isw` warns when the condition number exceeds \(10^6\).

```{r constant-head-diagnostics}
constant_head_diagnostics <- constant_head_schedule |>
  st_drop_geometry() |>
  summarize(
    maximum_absolute_boundary_residual_m = max(abs(as.numeric(
      set_units(boundary_residual, "m")
    ))),
    maximum_matrix_condition_number = max(matrix_condition_number),
    number_of_aquifer_factorizations = n_distinct(aquifer_id)
  )

constant_head_diagnostics
```

### Compare total stream depletion

Water lost from the stream enters the aquifer. The schedule reports this as a
negative `injection_rate` at each stream model point. Its negative is the
corresponding positive stream-depletion rate.

For ADF, this quantity is the interval-average analytical depletion distributed
among the segments. For constant head, it is the total injection needed to
maintain the boundary condition.

```{r plot-stream-depletion-comparison}
stream_depletion_comparison <- bind_rows(
  mutate(adf_schedule, method = "ADF"),
  mutate(constant_head_schedule, method = "Constant head")
) |>
  group_by(method, interval_end) |>
  summarize(
    stream_depletion_rate = -sum(injection_rate),
    .groups = "drop"
  )

ggplot(
  stream_depletion_comparison,
  aes(interval_end, stream_depletion_rate, color = method)
) +
  geom_step(linewidth = 1) +
  labs(
    x = "Interval end",
    y = "Total stream-depletion rate",
    subtitle = paste(
      "Positive stream depletion is represented internally as negative",
      "aquifer injection"
    ),
    color = "Method"
  ) +
  theme_minimal()
```

## Compare water levels at the stream

`get_aquifer_water_level_change()` can use either completed schedule. For the
constant-head schedule, net water-level change should be zero to numerical
precision at every collocation point and interval end. The method does not
enforce zero change between collocation points, so those values help reveal
how well the discrete wells represent the full stream boundary.

ADF calculates depletion but does not directly enforce water level at the
stream. Nonzero water-level changes are therefore expected for the ADF result.

```{r calculate-water-levels}
adf_water_levels <- get_aquifer_water_level_change(
  pumping_wells,
  pumping_schedules,
  observation_wells,
  stream_segments,
  evaluation_times,
  stream_injection_schedule = adf_schedule
) |>
  mutate(method = "ADF")

constant_head_water_levels <- get_aquifer_water_level_change(
  pumping_wells,
  pumping_schedules,
  observation_wells,
  stream_segments,
  evaluation_times,
  stream_injection_schedule = constant_head_schedule
) |>
  mutate(method = "Constant head")

water_levels <- bind_rows(adf_water_levels, constant_head_water_levels) |>
  left_join(st_drop_geometry(observation_wells), by = "observation_id") |>
  mutate(water_level_change_m = as.numeric(set_units(
    water_level_change,
    "m"
  )))
```

```{r plot-stream-boundary, fig.height = 6.5}
water_levels |>
  filter(
    observation_type == "stream boundary",
    evaluation_time == set_units(90, "days")
  ) |>
  ggplot(aes(
    as.numeric(distance_along_stream),
    water_level_change_m,
    color = method
  )) +
  geom_hline(yintercept = 0, color = "grey60") +
  geom_line(linewidth = 0.9) +
  geom_point(
    data = function(data) filter(data, is_collocation),
    aes(shape = "Collocation point"),
    size = 2
  ) +
  facet_wrap(~method, scales = "free_y", ncol = 1) +
  labs(
    x = "Distance along stream from pumping-well transect (m)",
    y = "Net water-level change (m)",
    color = "Method",
    shape = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r boundary-summary}
water_levels |>
  filter(observation_type == "stream boundary", is_collocation) |>
  group_by(method, evaluation_time) |>
  summarize(
    maximum_absolute_water_level_change_m = max(abs(
      water_level_change_m
    )),
    .groups = "drop"
  )
```

## Compare transects and the image-well solution

The transect plots show how the discrete stream representation behaves away
from the boundary. At day 90, the center transect can also be compared with
the analytical image-well solution for a straight, constant-head stream.

`isw` uses positive rates for pumping and negative rates for injection.
`get_straight_stream_drawdown_ratio()` returns the signed ratio
\(\Delta h / Q\), which is negative. Multiplying
the positive pumping rate directly by this ratio gives the signed water-level
change.

```{r image-well-reference}
image_well_reference <- tibble(
  method = "Image well",
  transect = "center",
  distance_from_stream = set_units(transect_distances, "m"),
  evaluation_time = set_units(90, "days"),
  water_level_change_m = as.numeric(set_units(
    set_units(400, "m^3/day") * get_straight_stream_drawdown_ratio(
      x1 = set_units(300, "m"),
      x2 = set_units(transect_distances, "m"),
      y_diff = set_units(0, "m"),
      K = pumping_wells$K[[1]],
      D = pumping_wells$D[[1]],
      V = pumping_wells$V[[1]],
      t = set_units(90, "days"),
      well_diam = pumping_wells$well_diam[[1]]
    ),
    "m"
  ))
)
```

```{r plot-transects, fig.height = 7}
transect_results <- water_levels |>
  filter(
    observation_type == "transect",
    evaluation_time == set_units(90, "days")
  ) |>
  select(method, transect, distance_from_stream, water_level_change_m)

bind_rows(transect_results, image_well_reference) |>
  ggplot(aes(
    as.numeric(distance_from_stream),
    water_level_change_m,
    color = method
  )) +
  geom_hline(yintercept = 0, color = "grey60") +
  geom_line(linewidth = 0.9, alpha = 0.65) +
  geom_point(size = 1.8, alpha = 0.65) +
  facet_wrap(~transect) +
  labs(
    x = "Distance from stream (m)",
    y = "Net water-level change (m)",
    color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Interpreting the comparison

The constant-head method used here represents the stream with discrete wells. The following checks help assess that
approximation:

1. boundary residuals at collocation points are near zero;
2. water levels between collocation points do not show unacceptable
   oscillation;
3. off-stream transects converge as segment and time spacing are refined; and
4. the response mimics the image-well solution for this straight,
   homogeneous test case.
