---
title: "Stream Segmentation and Drawdown at the Stream"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5.5,
  out.width = "100%",
  fig.align = "center"
)
```

```{r packages}
library(isw)
library(units)
library(tibble)
library(dplyr)
library(ggplot2)
library(sf)
```

## Purpose

The apportioned drawdown model represents stream depletion as injection at
one point per modeled stream segment. A zero-diameter point well has an
unbounded analytical response at zero distance and can therefore produce an
artificially large recovery value at an observation point on the stream.

`get_stream_segments()` regularizes this response by assigning each injection
well an effective diameter equal to half the segment's actual
`represented_length`. The response is consequently held constant inside a
radius of one quarter of that length. This analysis examines the remaining
sensitivity to stream-segment spacing:

* at several fixed observation points on the stream; and
* along perpendicular transects extending away from the stream.

The model reports water-level change, not absolute hydraulic head. Therefore,
`aquifer_drawdown = 0` is the relevant constant-head boundary condition at the
stream. The effective-diameter approach limits a numerical artifact but does
not enforce that boundary condition. A nonzero result at the stream is an ADF
boundary residual. The separate constant-head method solves the injection
rates as a coupled system; see the package vignette
`stream-injection-methods.Rmd` for that comparison.

## Construct a controlled straight-stream example

A straight stream makes distance from the boundary unambiguous. The pumping
well is 400 m east of the stream, and the 2 km stream is long enough to reduce
end effects near the observation transects. All coordinates are in meters in
a projected CRS.

```{r model-inputs}
analysis_crs <- 32615

pumping_wells <- st_as_sf(
  tibble(
    pump_id = "pump_1",
    x = 400,
    y = 0,
    K = set_units(10, "m/day"),
    D = set_units(30, "m"),
    V = 0.15,
    well_diam = set_units(0.3, "m")
  ),
  coords = c("x", "y"),
  crs = analysis_crs
)

stream_reaches <- st_sf(
  reach_id = "straight_stream",
  geometry = st_sfc(
    st_linestring(matrix(c(0, -1000, 0, 1000), ncol = 2, byrow = TRUE)),
    crs = analysis_crs
  )
)

pumping_schedules <- tibble(
  t = set_units(c(0, 180), "days"),
  pump_1 = set_units(c(500, 0), "m^3/day")
)
evaluation_times <- set_units(c(30, 90, 180), "days")
injection_times <- set_units(seq(10, 170, by = 10), "days")
```

## Place observations on the stream and along transects

The fixed on-stream locations do not depend on the segmentation. This avoids
moving the diagnostic points when segment spacing changes. Three transects
start at the stream and extend toward the pumping well. Their zero-distance
points are also on-stream observations.

```{r observation-grid}
on_stream <- tibble(
  observation_id = paste0("stream_y_", c("m500", "0", "p500")),
  transect = "on-stream checks",
  distance_from_stream = 0,
  x = 0,
  y = c(-500, 0, 500)
)

transects <- tidyr::expand_grid(
  transect_y = c(-500, 0, 500),
  distance_from_stream = c(0, 25, 50, 100, 200, 400, 800)
) |>
  mutate(
    observation_id = paste0("transect_", transect_y, "_x_", distance_from_stream),
    transect = paste0("y = ", transect_y, " m"),
    x = distance_from_stream,
    y = transect_y
  ) |>
  select(observation_id, transect, distance_from_stream, x, y)

observation_table <- bind_rows(on_stream, transects) |>
  distinct(observation_id, .keep_all = TRUE)

observation_wells <- st_as_sf(
  observation_table,
  coords = c("x", "y"),
  crs = analysis_crs
)

ggplot() +
  geom_sf(data = stream_reaches, color = "steelblue", linewidth = 1) +
  geom_sf(data = observation_wells, aes(color = transect), size = 1.8) +
  geom_sf(data = pumping_wells, color = "firebrick", size = 3) +
  coord_sf(xlim = c(-50, 850), ylim = c(-1050, 1050)) +
  labs(color = "Observation set") +
  theme_minimal()
```

## Run several stream-segment spacings

The target spacings below create progressively denser injection points.
`sample_spacing` is kept at one quarter of `reach_spacing` so the stream-web
apportionment is refined at the same time. The injection time grid is held
fixed across runs, separating spatial from temporal sensitivity.

```{r run-spacing-scenarios}
segment_spacings <- c(200, 100, 50, 20, 10)

run_spacing <- function(spacing_m) {
  stream_segments <- get_stream_segments(
    stream_reaches,
    reach_spacing = set_units(spacing_m, "m"),
    analysis_crs = analysis_crs
  )
  stream_apportionment <- get_adf_stream_apportionment(
    pumping_wells,
    stream_segments,
    sample_spacing = set_units(spacing_m / 4, "m"),
    method = "web_squared"
  )

  drawdown <- get_aquifer_water_level_change(
    pumping_wells,
    pumping_schedules,
    observation_wells,
    stream_segments,
    evaluation_times,
    injection_times = injection_times,
    injection_method = "adf",
    stream_apportionment = stream_apportionment
  )

  drawdown |>
    left_join(st_drop_geometry(observation_wells), by = "observation_id") |>
    mutate(
      segment_spacing_m = spacing_m,
      number_of_segments = n_distinct(stream_segments$reach_segment_id),
      aquifer_drawdown_m = -drop_units(set_units(
        water_level_change,
        "m"
      ))
    )
}

sensitivity_results <- bind_rows(lapply(segment_spacings, run_spacing))

sensitivity_results |>
  distinct(segment_spacing_m, number_of_segments)
```

## Inspect the stream boundary residual

Ideally, net drawdown is zero everywhere on a constant-head stream. This plot
shows the modeled residual at fixed stream locations. Comparing curves reveals
whether the result is converging as the stream representation is refined. A
logarithmic y-axis separates the small positive residuals; zero itself cannot
be displayed on this scale.

```{r plot-on-stream-residual}
sensitivity_results |>
  filter(transect == "on-stream checks") |>
  ggplot(aes(
    evaluation_time,
    aquifer_drawdown_m,
    color = factor(segment_spacing_m),
    group = segment_spacing_m
  )) +
  geom_line() +
  geom_point() +
  scale_y_log10() +
  facet_wrap(~observation_id) +
  labs(
    x = "Time since pumping began",
    y = "Net drawdown (m, log scale)",
    color = "Segment spacing (m)"
  ) +
  theme_minimal()
```

A compact diagnostic is the largest absolute on-stream residual for each
scenario and evaluation time.

```{r summarize-on-stream-residual}
stream_residual_summary <- sensitivity_results |>
  filter(transect == "on-stream checks") |>
  group_by(segment_spacing_m, number_of_segments, evaluation_time) |>
  summarize(
    maximum_absolute_residual_m = max(abs(aquifer_drawdown_m)),
    .groups = "drop"
  )

stream_residual_summary
```

## Inspect drawdown away from the stream

The transect profiles show how the modeled boundary residual relates to
drawdown within the aquifer. The final evaluation time is used here for a
readable spatial comparison. The pumping well lies at 400 m on the middle
transect, so its finite physical diameter also prevents a point singularity.
The logarithmic y-axis makes both the small near-stream response and larger
drawdown near the pumping well visible in the same figure.

```{r plot-transect-profiles}
final_time <- max(evaluation_times)

sensitivity_results |>
  filter(
    transect != "on-stream checks",
    evaluation_time == final_time
  ) |>
  ggplot(aes(
    distance_from_stream,
    aquifer_drawdown_m,
    color = factor(segment_spacing_m),
    group = segment_spacing_m
  )) +
  geom_line() +
  geom_point() +
  scale_y_log10() +
  facet_wrap(~transect) +
  labs(
    x = "Distance east of stream (m)",
    y = "Net drawdown (m, log scale)",
    color = "Segment spacing (m)"
  ) +
  theme_minimal()
```

## Interpretation

The effective injection diameter should be understood as a spatial
regularization tied to the length represented by each point. It makes values
at and near an injection point finite and makes the assumed support of the
point response change consistently with segment density.

The most useful convergence checks are:

1. whether the maximum absolute on-stream residual decreases or stabilizes as
   segment spacing is reduced; and
2. whether the off-stream transect profiles become insensitive to further
   refinement.

Neither result guarantees a constant-head boundary. Injection rates are still
assigned from apportioned stream depletion and are not solved simultaneously
to make net drawdown zero at all stream collocation points. Use
`method = "constant_head"` in `get_stream_injection_schedule()` when that
endpoint constraint is required. The residuals calculated here remain useful
for evaluating the ADF approximation.
